---
title: "honours_final_code_github"
output: html_notebook
---
# 1 - Pre-analysis
## 1.1 - Installing packages
```{r}
install.packages("dplyr")
install.packages("tidyr")
install.packages("readr")
install.packages("stringr")
install.packages("purrr")
install.packages("tibble")

install.packages("ggplot2")
install.packages("treemapify")
install.packages("patchwork")

install.packages("sf")
install.packages("ggspatial")
install.packages("prettymapr")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
install.packages("rmapshaper")

remotes::install_github("ropensci/rnaturalearthhires")
install.packages(
  "rnaturalearthhires",
  repos = "https://ropensci.r-universe.dev",
  type = "source"
)
install.packages("rnaturalearthhires")

install.packages("galah")

remotes::install_github("traitecoevo/austraits", dependencies = TRUE, upgrade = "ask")
remotes::install_github("traitecoevo/traits.build", dependencies = TRUE, upgrade = "ask")
```

## 1.2 Loading packages
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(purrr)
library(tibble)

library(ggplot2)
library(treemapify)
library(patchwork)

library(sf)
library(ggspatial)
library(prettymapr)
library(rnaturalearth)
library(rnaturalearthdata)
library(rnaturalearthhires)
library(rmapshaper)

library(galah)
library(austraits)    # functions for exploring a traits.build database, available at Github repo
library(traits.build) # additional functions for exploring a traits.build database, available at Github repo
```
## 1.3 Pulling AusTraits data
```{r}
source("https://raw.githubusercontent.com/traitecoevo/traits.build-book/master/data/extra_functions.R")

aus_traits <- austraits::get_versions() %>%
  dplyr::filter(version == "7.0.0") %>%
  dplyr::pull("doi")
aus_traits

austraits <- austraits::load_austraits(doi = aus_traits)
austraits
```

## 1.4 Data wrangling
### 1.4.1 importing and creating dataset
```{r}
# Importing data
setwd("/Users/dylan/Documents/GitHub/lost_flowers/honours_final_code_github")
master_data <- read.csv("data_searching_cleaned.csv")
master_data

# Filtering by searchable populations
searched_for_data <- master_data %>% 
  filter(refindable == "Y")
```
### 1.4.2 Cleaning dataset
```{r}
# Fixing taxonomic issues
searched_for_data <- searched_for_data %>%
  mutate(species_name = case_when(
    species_name == "Calamagrostis quadriseta" ~ "Deyeuxia quadriseta",
    species_name == "Sphaeropteris australis" ~ "Cyathea leichhardtiana",
    species_name == "Chenopodium trigonon" ~ "Einadia trigonos",
    TRUE ~ species_name))

# Adding a variable for binary found outcome
searched_for_data <- searched_for_data %>%
  mutate(found_simplified = case_when(
    found %in% c("Y", "?Y") ~"Yes",
    found %in% c("N", "?N") ~"No",
    found %in% c("?", "F", "") ~"?",
    TRUE ~ found)) %>%
  mutate(found_simplified_withF = case_when(
    found %in% c("Y", "?Y") ~"Yes",
    found %in% c("N", "?N") ~"No",
    found %in% c("?", "") ~"?",
    found %in% c("F") ~ "Failed search",
    TRUE ~ found)) %>%
  mutate(found_simplified_number = case_when(
    found_simplified %in% c("Yes") ~1,
    found_simplified %in% c("No") ~0,
    found_simplified %in% c("?") ~NA))

# To be able to choose the order of these response variables when graphing
searched_for_data$found_factor <- factor(searched_for_data$found, levels = c("Y", "?Y", "N", "?N", "F", "?", "")) 

# Splitting lat and long into separate columns for mapping
searched_for_data <-
  searched_for_data %>%
  separate(search_area_coordinates, into = c("search_area_lat", "search_area_long"), sep = ", ", convert = TRUE, remove = FALSE)
```
### 1.4.3 Filtering AusTraits data
```{r}
# creating a subset of Austraits data with only target species
austraits_searching <- 
  austraits$traits %>%
  filter(taxon_name %in% c(searched_for_data$species_name)) %>% # filter by only for species searched for
  left_join(searched_for_data %>% # add re-found or not column
              select(species_name, found), by = c("taxon_name" = "species_name")) %>%
  mutate(found_simplified = case_when(
    found %in% c("Y", "?Y") ~"Yes",
    found %in% c("N", "?N") ~"No",
    found %in% c("?", "F", "") ~"?",
    TRUE ~ found)) %>%
  mutate(found_simplified_number = case_when(
    found_simplified %in% c("Yes") ~1,
    found_simplified %in% c("No") ~0,
    found_simplified %in% c("?") ~NA))

# filtering out unknowns
austraits_searching_foundYorN <- 
  austraits_searching %>%
  filter(found_simplified != "?")

# adding woodiness data
woodiness_data <- austraits_searching %>%
  filter(trait_name == "woodiness_detailed")

# filtering out conflicting values
woodiness_data_cleaned <- 
  woodiness_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) == 1) %>% # filtering out species with conflicting values
  distinct(taxon_name, .keep_all = TRUE) %>%
  ungroup()

#filtering out "?" for graphs and stats - didnt do originally as adding woodiness data to main dataset, to wanted to have it for all
woodiness_data_cleaned <- 
  woodiness_data_cleaned %>%
  filter(found_simplified != "?")

#ADDING WOODINESS DATA TO MAIN DATA TABLE
searched_for_data <- 
  searched_for_data %>%
  select(-any_of("woodiness")) %>% #removing column if it already exists (to avoid duplicates)
  left_join(
    woodiness_data_cleaned %>%
      select(taxon_name, value), by = c("species_name" = "taxon_name")) %>%
  rename(woodiness = value) # renaming column to "woodiness"

#removing "?"
searched_for_data_YorN <- searched_for_data %>%
  filter(found_simplified != "?")
```

### 1.4.4 Checking stuff
```{r}
n_distinct(austraits_searching_foundYorN$taxon_name) # checking number of species that I've filtered to
setdiff(searched_for_data$species_name, austraits_searching_foundYorN$taxon_name) # checking missing species - should be good now (the 7 missing ones had "?" found value)

# searching a row in data frame
searched_for_data %>%
  filter(species_name == "Flagellaria indica") %>% # change species name and / or trait name
  select(species_name, woodiness) # what columns i want to view

austraits::lookup_trait(database = austraits, term = "dispersal") # use to look up traits: place search term after term = "..."

filter(austraits$traits,
       taxon_name %in% c("Gonocarpus tetragynus"),
       trait_name == "woodiness_detailed") # use to look up species in AusTraits

austraits %>%  # check values of a given trait
  pluck("traits") %>%
  filter(trait_name == "dispersal_appendage") %>% #CHANGE TRAIT NAME HERE
  pull("value") %>% 
  unique()

# checking which species are missing values
austraits_searching_foundYorN %>%
  distinct(taxon_name) %>%
  anti_join(
    austraits_searching_foundYorN %>%
      filter(trait_name == "seedbank_longevity") %>%
      distinct(taxon_name),
    by = "taxon_name"
  )

# checking what species have a given trait entry
austraits_searching_foundYorN %>%
  filter(trait_name == "serotiny") %>% # enter trait here
  arrange(taxon_name)
```



# 2 - Trait analysis
## 2.1 - Height
### 2.1.1 - Height data wrangling
```{r}
# Extract height data
height_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "plant_height", 
         life_stage != "seedling") %>% # Filtering out data points from seedlings
  mutate(value=as.numeric(value)) # changing plant height data to numerical

# NOTE: there are also min and mean values, but not using these
max_height_data <- height_data %>% 
  filter(value_type == "maximum")

# Adding a column with mean max plant height to data table
max_height_data_cleaned <- max_height_data %>% 
  group_by(taxon_name) %>%
  mutate(average_height = mean(value, na.rm = TRUE)) %>%
  distinct(taxon_name, .keep_all = TRUE)
```
### 2.1.2 - Height violin plots
```{r}
log_height_plot <- 
  ggplot(max_height_data_cleaned, aes(x = average_height, y = found_simplified, fill = found_simplified)) +
  geom_violin() +
  theme_minimal() +
  labs(
    title = "Re-findability by plant height (n = 77)",
    x = "Average height (log transformed m)", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_log10() +
  theme(legend.position = "none")
log_height_plot
```
### 2.1.3 - Height stats
```{r}
summary(
  max_height_data_cleaned %>%
  glm(as.numeric(found_simplified_number) ~ log10(average_height),
      family = binomial(link = "logit"),
      data = .)
  )
```

## 2.2 Woodiness
### 2.2.1 - Woodiness data wrangling
```{r}
# NOTE - some species have different values depending on source, when filtering, the most common is selected.
#   Eleocharis sphacelata: “herbaceous woody_root”(1), “herbaceous” (5)
#   Gonocarpus tetragynus: “herbaceous woody_base” (1), “herbaceous” (4)
#   Flagellaria indica: "herbaceous" (2), "woody" (1)
#   Monoxtaxis linifolia: "woody base" (1), "herbaceous woody base" (1)
#   Piper hederaceum: "herbaceous" (1), "herbaceous woody" (1)
#   Opercularia aspera: "herbaceous" (1), "herbaceous woody" (1)
#   Senecio pinnatifolius: “herbaceous semi-woody” (1), “herbaceous" (many)

#checking species with conflicting trait values 
woodiness_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) > 1) %>% #filtering for multiple trait values
  count(taxon_name, value) %>% #adding a count column
  arrange(taxon_name) # sorting by species


table(woodiness_data_cleaned$value) # checking
```
### 2.2.2 - Woodiness graph
```{r}
woodiness_plot <- 
  ggplot(woodiness_data_cleaned, aes(x = value, fill = found_simplified)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Re-findability by woodiness (n = 88)",
    x = "Woodiness", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_discrete(
    labels = c("herbaceous" = "Herbaceous", "woody" = "Woody"))
woodiness_plot
```
### 2.2.3 - Woodiness stats
```{r}
woodiness_table <- table(woodiness_data_cleaned$value, woodiness_data_cleaned$found_simplified) # creating a contingency table
chisq.test(woodiness_table) # chi squared test
chisq.test(woodiness_table)$residuals # gives a value for each value interaction
```
## 2.3 - Resprouting capability
### 2.3.1 - Resprouting data wrangling
```{r}
resprouting_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "resprouting_capacity") %>%
  arrange(taxon_name)

# Assigning the value "partial_resprouting" to species with conflicting data
resprouting_data_cleaned <- resprouting_data %>%
  group_by(taxon_name) %>%
  mutate(value = if (n_distinct(value) > 1) "partial_resprouting" else unique(value)) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(taxon_name)
```
### 2.3.2 - Resprouting graph
```{r}
resprouting_plot <-
  ggplot(resprouting_data_cleaned, aes(x = value, fill = found_simplified)) +
  geom_bar(position = "stack") +
  theme_minimal() +
  labs(
    title = "Re-findability by resprouting capability \n(n = 80)",
    x = "Resprouting capability", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
scale_x_discrete(
  labels = c("fire_killed" = "Killed by fire", "partial_resprouting" = "Partial\nresprouting", "resprouts" = "Resprouter"))
resprouting_plot
```

### 2.3.3 - Resprouting stats
```{r}
resprouting_table <- table(resprouting_data_cleaned$found_simplified, resprouting_data_cleaned$value)
chisq.test(resprouting_table)
chisq.test(resprouting_table)$residuals
```
## 2.4 Seedbank
### 2.4.1 Seedbank data wrangling
```{r}
seedbank_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "seedbank_location") %>%
  mutate(value_simplified = case_when(
    value %in% c("soil_seedbank", "canopy_seedbank") ~"seedbank",
    TRUE ~ value)) %>%
  arrange(taxon_name)

# Removing species with conflicting values
seedbank_data_cleaned <- 
  seedbank_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) == 1) %>%
  slice(1) %>%
  arrange(taxon_name)
```
### 2.4.2 Seedbank graph
```{r}
seedbank_simplified_plot <-
  ggplot(seedbank_data_cleaned, aes(x = value_simplified, fill = found_simplified)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Re-findability by seedbank presence (n = 33)",
    x = "Seedbank presence", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_discrete(
    labels = c("seedbank" = "Present", "none" = "Absent"))
seedbank_simplified_plot
```
### 2.4.3 - Seedbank stats
```{r}
seedbank_simplified_table <- table(seedbank_data_cleaned$found_simplified, seedbank_data_cleaned$value_simplified)
chisq.test(seedbank_simplified_table)
chisq.test(seedbank_simplified_table)$residuals
```

## 2.5 - Latitude quantile
### 2.5.1 - Galah data wrangling
```{r}
galah_config(atlas = "ALA",
             email = "d.petrozzi@student.unsw.edu.au")

# Creating data frame with all target species' ALA records (only need to run once)
ALA_data <-
  galah_call() %>%
  identify(searched_for_data$species_name) %>%
  collect()

# Pulling a list of species names as they appear in ALA
ALA_taxa_data <- search_taxa(searched_for_data$species_name) %>%
  select(search_term, scientific_name) %>% # choosing columns to remain
  distinct(search_term, scientific_name) %>%
  collect()

# And then joining to existing data
searched_for_data <- 
  searched_for_data %>%
  select(-any_of("scientific_name_ALA")) %>% #removing column if it already exists (to avoid duplicates)
  left_join(ALA_taxa_data, by = c("species_name" = "search_term")) %>%
  rename(scientific_name_ALA = scientific_name) # renaming column to scientific_name_ALA


# Adding range quantile column
searched_for_data <-
  searched_for_data %>%
  select(-any_of("latitude_percentile")) %>% # Removing column if it's already there to avoid duplication
  left_join(
    ALA_data %>%
      filter(occurrenceStatus == "PRESENT") %>%
      filter(!is.na(decimalLatitude)) %>% # removing NAs
      group_by(scientificName) %>%
      summarise(latitude_percentile = ecdf(decimalLatitude)(-34.13)*100), # empirical cumulative distribution functions (ecdf)
    by = c("scientific_name_ALA" = "scientificName")
  )
```
### 2.5.2 Latitude quantile graph and stats
```{r}
range_plot <- searched_for_data %>%
  filter(found_simplified_number != "?") %>%
  ggplot(aes(x = latitude_percentile, y = found_simplified, fill = found_simplified)) +
  geom_violin() +
  theme_minimal() +
  labs(
    title = "Re-findability by latitude quantile (n = 95)",
    x = "Latitude quantile", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  theme(legend.position = "none") +
  scale_x_continuous(
    breaks = c(0, 25, 50, 75, 100),
    labels = c("0\n(South\nrange\nedge)", "25", "50", "75", "100\n(North\nrange\nedge)"))
range_plot

glm(as.numeric(found_simplified_number) ~ latitude_percentile,
    family = binomial(link = "logit"),
    data = searched_for_data) %>% # work our why na.omit doesn't work
  summary()
```
# 3 - Methods analysis
## 3.1 - Search type
```{r}
individual_or_group_search_plot <- ggplot(searched_for_data_YorN, aes(x = individual_or_multiple_search, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by search type (n = 95)",
    x = "Was the target species searched for\nexclusively or alongside others?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("I" = "Exclusively", "M" = "Alongside others"))
individual_or_group_search_plot

individual_or_multiple_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$individual_or_multiple_search)
chisq.test(individual_or_multiple_table)
chisq.test(individual_or_multiple_table)$residuals
```
## 3.2 - Species familiarity
```{r}
seen_before_plot <- 
  ggplot(searched_for_data_YorN, aes(x = volunteers_seen_before, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by species familiarity (n = 95)",
    x = "Had the searchers seen\nthe species before?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("N" = "No", "Y" = "Yes"))
seen_before_plot

volunteers_seen_before_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$volunteers_seen_before)
chisq.test(volunteers_seen_before_table)
chisq.test(volunteers_seen_before_table)$residuals
```


## 3.3 - Uniqueness
### 3.3.1 - Uniqueness data wrangling
```{r}
royal_master_list <- 
  read.csv("royal_master_list.csv") %>%
  mutate(genus = word(accepted_name, 1)) # adding genera column

unique_genera_data <- 
  royal_master_list %>%
  group_by(genus) %>%
  filter (n() == 1) %>%
  ungroup() %>%
  filter(accepted_name %in% searched_for_data$species_name)

searched_for_data <-
  searched_for_data %>%
  mutate(unique_genera_in_royal = species_name %in% unique_genera_data$accepted_name)
```
### 3.3.2 - Uniqueness graph and stats
```{r}
unique_in_royal_plot <- 
  searched_for_data %>%
  filter(found_simplified != "?") %>%
  ggplot(aes(x = unique_genera_in_royal, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")
  ) +
  labs(
    title = "Re-findability by uniqueness (n = 95)",
    x = "Is the species the only one\nof its genera in Royal?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(labels = c("FALSE" = "No", "TRUE" = "Yes"))
unique_in_royal_plot

unique_in_royal_table <- searched_for_data %>%
  filter(found_simplified != "?") %>%
  xtabs(~ found_simplified + unique_genera_in_royal, data = .)
chisq.test(unique_in_royal_table)
chisq.test(unique_in_royal_table)$residuals
```



## 3.4 - Associated species
```{r}
associated_species_plot <- 
  ggplot(searched_for_data_YorN, aes(x = associated_species_known, fill = found_simplified)) + 
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by associated species\n(n = 95)",
    x = "Were associated species known?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("N" = "No", "Y" = "Yes"))
associated_species_plot

associated_species_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$associated_species_known)
chisq.test(associated_species_table)
chisq.test(associated_species_table)$residuals
```
## 3.5 - Search shape
```{r}
linear_or_area_plot <- 
  ggplot(searched_for_data_YorN, aes(x = linear_or_area_search, fill = found_simplified)) + 
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by search shape (n = 95)",
    x = "Was a linear or area-based\nsearch conducted?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(labels = c("A" = "Area", "L" = "Linear"))
linear_or_area_plot

linear_or_area_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$linear_or_area_search)
chisq.test(linear_or_area_table)
chisq.test(linear_or_area_table)$residuals
```

## 3.6 Flowering
```{r}
flowering_time_plot <- 
  searched_for_data_YorN %>%
  filter(searched_in_flowering_time != "#N/A") %>%
   ggplot(aes(x = searched_in_flowering_time, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by search timing (n = 71)",
    x = "Was search conducted during\nflowering time?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("N" = "No", "Y" = "Yes"))
flowering_time_plot

flowering_time_table <- 
  searched_for_data_YorN %>%
  filter(searched_in_flowering_time != "#N/A") %>%
  with(table(found_simplified, searched_in_flowering_time))
chisq.test(flowering_time_table)
chisq.test(flowering_time_table)$residuals

# Species found with fertile material outside flowering time
searched_for_data %>%
  filter(searched_in_flowering_time == "N", 
         fertile_material_present == "Y",
         found_simplified == "Yes") %>%
  select(species_name, searched_in_flowering_time, fertile_material_present, found)
```


# 4 - Mapping
## 4.1 - Royal NP map and data wrangling
### 4.1.1 - Royal NP map and outline
```{r}
Royal_map <- st_read("royal_map/Royal_outline.shp") %>%
  st_transform(crs = 4326) %>% # assigning WGS 84 coordinate system
  st_make_valid()

Royal_roads <- st_read("royal_map/Royal_roads.gpx", layer = "tracks") %>% # reading road file
  st_transform(crs = 4326) %>%
  st_make_valid() %>%
  { rmapshaper::ms_clip(target = ., clip = Royal_map[1]) } # limiting extent of roads to borders of Royal map

searched_for_data_sf <- searched_for_data %>%
  st_as_sf(coords = c("search_area_long", "search_area_lat"), crs = 4326) %>%
  st_transform(crs = st_crs(Royal_map))
```
### 4.1.2 - Royal NP long unobserved species
```{r}
royal_all_long_unobserved <- read.csv("royal_long_unobserved_all.csv")

royal_all_long_unobserved_sf <- royal_all_long_unobserved %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = st_crs(Royal_map))
```
### 4.1.3 - Australia map inset
```{r}
royal_bbox <- st_as_sfc(st_bbox(Royal_map))

aus_map <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf") # Australia map
aus_states <- ne_states(country = "Australia", returnclass = "sf") # state boundaries
royal_coords <- st_sf(
              name = "Royal National Park", 
              geometry = st_sfc(st_point(c(151.07, -34.13))), 
              crs = 4326)

aus_map_simple <- ne_countries(scale = "small", country = "Australia", returnclass = "sf")

inset_aus_map <- inset_element(ggplot() +
  geom_sf(data = aus_map_simple, fill = "#D8E7D0", color = "grey50") +
  geom_sf(data = royal_coords, color = "red", size = 1, shape = 0, stroke = 0.7) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.3)
  ), 
    left = 0.7,   # horizontal placement (0 = left, 1 = right)
    bottom = -0.4, # vertical placement (0 = bottom, 1 = top)
    right = 0.98,  # width of inset area
    top = 0.98     # height of inset area
)
```
## 4.2 - Map building
### 4.2.1 Native long unobserved by year
```{r}
# fig 1
long_unobserved_native_by_year_map <-
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = subset(royal_all_long_unobserved_sf, establishment_means == "native"), aes(color = year), size = 1, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  +
   scale_color_gradientn(
    name = "Year last recorded",
    colors = c("yellow", "orange", "purple4"),
    breaks = c(seq(1870, 1990, by = 20)),
    limits = c(1870, 1990),
    na.value = "purple4",
    guide = guide_colourbar(
    barheight = unit(7, "cm"),
    barwidth = unit(0.5, "cm"))) +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  ) +
  labs(
    colour = "Establishment means"
  )
long_unobserved_native_by_year_map + inset_aus_map
```
### 4.2.2 Target species by search outcome
```{r}
searched_map_by_found <- 
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = searched_for_data_sf, aes(color = found_simplified_withF), size = 1.5, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  +
   scale_color_manual(
    name = "Search outcome",
    values = c(
    "Yes" = "orange",
    "No" = "darkblue",
    "?" = "grey",
    "Failed search" = "brown"),
    labels = c(
      "?" = "Unknown (1)",
      "Yes" = "Found (27)",
      "No" = "Not found (68)",
      "Failed search" = "Search not conducted (4)"
    )) +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  )
searched_map_by_found + inset_aus_map
```
## 4.3 - Range edge maps
### 4.3.1 - Range data wrangling
```{r}
# Checking range edge species, selecting top 3 and bottom 3 to visualize
searched_for_data %>%
  arrange(latitude_percentile) %>%
  select(species_name, latitude_percentile)
```
### 4.3.2 Range map creation
```{r}
species_range_map_function <- function(species) {
  
  coords <- 
    galah_call() %>%
    identify(species) %>%
    geolocate(aus_map_simple) %>%
    collect() %>%
    filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
    st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
  
  ggplot(data = aus_map) +
    geom_sf(fill = "#C5DDC2", 
            color = "grey50", 
            alpha = 0.8) +
    geom_sf(data = aus_states, 
            fill = NA, 
            color = "gray30") +
    geom_sf(data = coords, 
            aes(color = "Observations"), 
            size = 1.3) +
    geom_sf(data = royal_coords, 
            aes(color = "Royal National Park"), 
            size = 2) +
    scale_color_manual(
      values = c(
        "Royal National Park" = "blue4", 
        "Observations" = "darkorange"),
      name = NULL) +
    coord_sf(
      xlim = c(110, 157),
      ylim = c(-45, -9),
      expand = FALSE
    ) + 
    theme_minimal() +
    labs(title = species)
}


range_extreme_species_maps <- map(c(
  "Juncus pallidus", 
  "Leptospermum continentale", 
  "Gahnia radula",
  "Flagellaria indica", 
  "Cyperus cyperoides", 
  "Sacciolepis indica"
), species_range_map_function)
```

### 4.3.3 - Building range map figure
```{r}
# Creating a map theme
range_map_theme <- 
  function(m) { 
    m + theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "italic", , size = 9))}

range_edge_species_maps <- 
 range_extreme_species_maps %>%
  lapply(range_map_theme) %>%
  wrap_plots(ncol = 3, guides = "collect") + 
  plot_annotation(tag_levels = 'a') &
  theme(legend.position = "bottom")
range_edge_species_maps
```

# 5 - Other graphs
## 5.1 - Treemap
### 5.1.1 - Treemap data wrangling
```{r}
#fixing names as they differ on ALA
master_data <- 
  master_data %>%
  mutate(species_name_ALA = case_when(
    species_name == "Calamagrostis quadriseta" ~ "Deyeuxia quadriseta",
    species_name == "Sphaeropteris australis" ~ "Cyathea leichhardtiana",
    species_name == "Chenopodium trigonon" ~ "Einadia trigonos",
    species_name == "Cyanothamnus rigens" ~ "Boronia rigens",
    species_name == "Setaria distans" ~ "Paspalidium distans",
    species_name == "Microsorum pustulatum" ~ "Zealandia pustulata", # multiple synonyms, but the record links to this species
    species_name == "Genoplesium nudiscapum" ~ "Corunastylis nudiscapa",
    species_name == "Genoplesium woollsii" ~ "Corunastylis woollsii",
    species_name == "Lastreopsis decomposita" ~ "Parapolystichum calanthum", # AusTraits uses this name for the one trait record, as opposed to the accepted Lastreopsis calantha
    species_name == "Stolonochloa pygmaea" ~ "Panicum pygmaeum",
    species_name == "Zostera capricorni" ~ "Zostera muelleri subsp. capricorni",
    species_name == "Gaudium parvifolium" ~ "Leptospermum parvifolium",
    TRUE ~ species_name))

# Attaching AusTraits data to long unobserved data
austraits_long_unobserved <- 
  austraits$traits %>%
  filter(taxon_name %in% c(master_data$species_name)) %>%
  left_join(master_data %>% # add re-found or not column
              select(species_name, taxonomy_issue, location_issue, refindable), by = c("taxon_name" = "species_name"))
# And then to woodiness
woodiness_long_unobserved <-
  austraits_long_unobserved %>%
  filter(trait_name == "woodiness_detailed") %>%
  filter(value %in% c("woody", "herbaceous")) %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) == 1) %>%
  ungroup() %>%
  distinct(taxon_name, .keep_all = TRUE) %>%
  rename(woodiness = value) %>% # renaming column to "woodiness" 
  select(taxon_name, woodiness, taxonomy_issue, location_issue) %>%
  mutate(refindable = case_when(
    location_issue == "Y" & taxonomy_issue == "Y" ~ "Location and taxonomy issue",
    location_issue == "Y" & taxonomy_issue == "N" ~ "Location issue",
    location_issue == "N" & taxonomy_issue == "Y" ~ "Taxonomy issue",
    location_issue == "N" & taxonomy_issue == "N" ~ "No issues"
  ))
```
### 5.1.2 - Treemap graph 
```{r}
woodiness_treemap <- 
  woodiness_long_unobserved %>%
  count(woodiness, refindable) %>%
  mutate(woodiness = factor(woodiness))

curation_treemap <- 
  ggplot(woodiness_treemap, 
         aes(area = n, 
             fill = refindable, 
             subgroup = woodiness, 
             label = n)) +
    geom_treemap(color = "black", size = 2) +  
    geom_treemap_subgroup_border(color = "black", size = 2) + 
    geom_treemap_text(
      aes(label = n),
      place = "centre",
      size = 8,
      colour = "white",
      grow = FALSE,
      fontface = "bold"
    ) +
    geom_treemap_subgroup_text(
    place = "centre",
    colour = "black", 
    size = 12, 
    fontface = "bold"
    ) +
    labs(fill = "Re-findable?") +
    theme(legend.position = "right") +
  scale_fill_manual(
    values = c("Location and taxonomy issue" = "darkblue",
               "Location issue" = "grey",
               "Taxonomy issue" = "brown",
               "No issues" = "orange"),
    labels = c("Location and taxonomy issue" = "No (Location and taxonomy issue)",
               "Location issue" = "No (Location issue)",
               "Taxonomy issue" = "No (Taxonomy issue)",
               "No issues" = "Yes"),
     breaks = c("No issues", "Taxonomy issue", "Location issue", "Location and taxonomy issue"))

curation_treemap
```
## 5.2 - Long unobserved by family
### 5.2.1 - Family data wrangling
```{r}
family_all_species_data <-
  master_data %>%
    count(family) %>%
    mutate(family = if_else(n < 5, "Other families", family)) %>%
    group_by(family) %>% 
    summarise(n = sum(n), .groups = "drop") %>%
    mutate(family = reorder(family, n)) %>%
    mutate(life_form = case_when(
      family %in% c("Cyperaceae", "Poaceae", "Juncaceae", "Restionaceae") ~ "Graminoids",
      family %in% c("Myrtaceae", "Fabaceae", "Asteraceae", "Rhamnaceae", "Proteaceae", "Polygonaceae",
                      "Plantaginaceae", "Lauraceae", "Ericaceae", "Araliaceae") ~ "Eudicots",
      family %in% c("Orchidaceae", "Asparagaceae") ~ "Non-graminoid monocots",
      TRUE ~ "Other families"
    ))
```

### 5.2.2 - Family plot
```{r}
family_all_species_plot <- 
  ggplot() + 
    geom_bar(data = filter(family_all_species_data, family != "Other families"),
             aes(x = family, y = n, fill = life_form),
             stat = "identity") +
    geom_bar(data = filter(family_all_species_data, family == "Other families"),
             aes(x = family, y = n, fill = life_form),
             stat = "identity",
             show.legend = FALSE) +
    coord_flip() +
    geom_text(
      data = family_all_species_data,
      aes(x = family, y = n, label = n),
      hjust = -0.5, 
      size = 3, 
      fontface = "bold") +
    theme_classic() + 
    theme() +
    scale_fill_manual(
      values = c("Graminoids" = "brown",
                 "Eudicots" = "orange",
                 "Non-graminoid monocots" = "darkblue"
    )) +
    labs(y = "Species count",
         fill = "Life form",
         x = "Family") +
  scale_y_continuous(
    limits = c(0, 120)
  )
family_all_species_plot
```
## 5.3 - Abundance
### 5.3.1 - Abundance data wrangling
```{r}
abundance_data <- searched_for_data %>%
  mutate(individuals_found_cleaned = case_when(
         individuals_found %in% c("") ~ NA,
         TRUE ~ individuals_found),
         individuals_found_cleaned = parse_number(individuals_found_cleaned)) %>% # pulling only numbers
  mutate(growth_form = case_when(
    quadrat_size_m2 %in% c(64) ~ "Herb",
    quadrat_size_m2 %in% c(144) ~ "Shrub",
    quadrat_size_m2 %in% c(400) ~ "Tree")) %>%
  filter(found_simplified == "Yes") %>%
  mutate(abundance_density = individuals_found_cleaned / quadrat_size_m2,
        abundance_density = case_when(
      individuals_found == "many" ~ 0.7, # manually assign abundance density to "many"
      TRUE ~ abundance_density)) %>%
  mutate(abundance_density = as.numeric(abundance_density)) %>%
  select(species_name, individuals_found, individuals_found_cleaned, found_simplified, quadrat_size_m2, abundance_density, growth_form)
```
### 5.3.2 - Abundance data plot and stat
```{r}
abundance_density_plot <- 
  ggplot(
    filter(abundance_data, !is.na(abundance_density)), 
    aes(x = abundance_density*10, fill = growth_form)) +
  geom_histogram(binwidth = 1, 
                 color = "black",
                 linewidth = 0.4) +
  geom_text(
    data = filter(abundance_data, !is.na(abundance_density)),
    aes(x = abundance_density * 10, label = ifelse(..count.. == 0, "", ..count..)),
    stat = "bin",
    binwidth = 1,
    color = "black",
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  theme_classic() +
  labs(
    x = bquote("Aboveground stems per 10m"^2),
    y = "Species",
    fill = "Growth form") +
  scale_x_continuous(breaks = seq(0, 7, by = 1),
                     labels = c("<1", "1 - <2", "2 - <3", "3 - <4", "4 - <5", "5 - <6", "6 - <7", "≥7")) +
  scale_fill_manual(values = c("Herb" = "brown", "Shrub" = "orange", "Tree" = "darkblue"))
abundance_density_plot
```
## 5.4 - Example search path
```{r}
gaia_gpx <- st_read("sphaeropteris-australis.gpx", layer = "tracks")

# Transform to a common CRS (Web Mercator for OSM
Royal_map_osm <- st_transform(Royal_map, 3857)
Royal_roads_osm <- st_transform(Royal_roads, 3857)
gaia_gpx_osm <- st_transform(gaia_gpx, 3857)

# Compute bounding boxes
gaia_bbox_numeric <- st_bbox(gaia_gpx_osm)
gaia_bbox_osm <- st_as_sfc(gaia_bbox_numeric)
Royal_bbox <- st_bbox(Royal_map_osm)

# Main GPX map
gaia_path_map <- ggplot() +
  # Base map
  annotation_map_tile(type = "osm", zoom = 18) +
  
  # Polygon map and roads
  geom_sf(data = Royal_map_osm, fill = NA, color = "grey50", alpha = 0.8) +
  geom_sf(data = Royal_roads_osm, color = "grey30") +
  
  # GPX track
  geom_sf(data = gaia_gpx_osm, color = "red", linewidth = 0.7) +
  
  # Scale and north arrow
  annotation_scale(location = "bl", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm")) +
  annotation_north_arrow(location = "tr", which_north = "true", style = north_arrow_fancy_orienteering(),
                         height = unit(1, "cm"), width = unit(1, "cm"), pad_x = unit(0.5, "cm"), pad_y = unit(0.5, "cm")) +
  
  # Coordinate limits around GPX with buffer
  coord_sf(
    xlim = c(gaia_bbox_numeric["xmin"], gaia_bbox_numeric["xmax"]),
    ylim = c(gaia_bbox_numeric["ymin"], gaia_bbox_numeric["ymax"]),
    expand = TRUE
  ) +
  
  # Clean theme
  theme_classic() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    plot.margin = unit(c(0,0,0,0), "cm")
  )

# Inset Royal map
inset_royal_map <- inset_element(
  ggplot() +
    geom_sf(data = Royal_map_osm, fill = "#D8E7D0", color = "grey50") +
    geom_sf(data = gaia_bbox_osm, color = "red", linewidth = 0.9) +
    coord_sf(
      xlim = c(Royal_bbox["xmin"], Royal_bbox["xmax"]),
      ylim = c(Royal_bbox["ymin"], Royal_bbox["ymax"]),
      expand = TRUE
    ) +
    theme_void() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_rect(color = "black", fill = NA)
    ),
  left = 0.55, bottom = 0.35, right = 0.95, top = 1
)

# Combine main map + inset
gaia_final_map <- gaia_path_map + inset_royal_map
gaia_final_map
```
## 5.5 - Re-findability by year
```{r}
year_plot_YorN <- 
  ggplot(searched_for_data_YorN, aes (x=sampled_year, fill = found_simplified_withF)) + # bar plot with samples collected by year
    geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    x = "Year",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(searched_for_data$sampled_year), max(searched_for_data$sampled_year), by = 2),
                     labels = function(x) ifelse(x %% 4 == 0, x, ""))
year_plot_YorN
```
# 6 - Printing final plots
## 6.1 - Plot data wrangling
```{r}
# creating a theme for figure format
figure_theme <- (function(p) { 
    p + theme(legend.position = "none", 
              plot.title = element_text(face = "bold", size = 10))})

#combining plots

trait_plots <-
  list(
  woodiness_plot,
    resprouting_plot,
    seedbank_simplified_plot,
    log_height_plot, 
    range_plot
    ) %>%
    lapply(figure_theme) %>%
    wrap_plots(ncol = 2, guides = "collect") +
    plot_annotation(tag_levels = 'a') +
    theme(legend.position = "right")

methodology_plots <-
  list( 
    individual_or_group_search_plot, 
    seen_before_plot,
    unique_in_royal_plot,
    associated_species_plot, 
    linear_or_area_plot,
    flowering_time_plot
    ) %>%
    lapply(figure_theme) %>%
    wrap_plots(ncol = 2, guides = "collect") + # guide refers to things like legends
    plot_annotation(tag_levels = 'a') +  #adding labels a, b, c...
    theme(legend.position = "right")     # Set overall legend position
```

## 6.2 - Methodology plots
```{r}
long_unobserved_native_by_year_map
fig_1 <- long_unobserved_native_by_year_map + inset_aus_map # fig 1
  ggsave("plots/fig_1.png", plot = fig_1, width = 6, height = 6, dpi = 300)

family_all_species_plot # fig 2
  ggsave("plots/fig_2.png", plot = family_all_species_plot, width = 8, height = 6, dpi = 300)

range_edge_species_maps # fig 3
  ggsave("plots/fig_3.png", plot = range_edge_species_maps, width = 8, height = 6, dpi = 300)

curation_treemap # fig S1
  ggsave("plots/fig_S1.png", plot = curation_treemap, width = 8, height = 5, dpi = 300)
```
## 6.3 - Results plots
```{r}
searched_map_by_year
searched_map_by_found
fig_6 <-searched_map_by_found + inset_aus_map # fig 6
  ggsave("plots/fig_6.png", plot = fig_6, width = 6, height = 6, dpi = 300)

year_plot_YorN # fig 7
  ggsave("plots/fig_7.png", plot = year_plot_YorN, width = 8, height = 6, dpi = 300)
year_plot
decade_plot

trait_plots # fig 8
  ggsave("plots/fig_8.png", plot = trait_plots, width = 8, height = 6, dpi = 300)
methodology_plots # fig 9
  ggsave("plots/fig_9.png", plot = methodology_plots, width = 8, height = 6, dpi = 300)

abundance_density_plot # fig 10
  ggsave("plots/fig_10.png", plot = abundance_density_plot, width = 8, height = 6, dpi = 300)

gaia_final_map # fig 11
  ggsave("plots/fig_11.png", plot = gaia_final_map, width = 3, height = 5, dpi = 300)
```