---
title: "trait_analysis_honours"
output: html_document
date: "2025-08-14"
---
# importing packages
**importing packages, AusTraits**
```{r}
#importing necessary AusTraits packages

install.packages("patchwork")
install.packages("galah")
remotes::install_github("traitecoevo/austraits", dependencies = TRUE, upgrade = "ask")
remotes::install_github("traitecoevo/traits.build", dependencies = TRUE, upgrade = "ask")

install.packages("treemapify")

install.packages("rnaturalearth")
install.packages("rnaturalearthdata")

remotes::install_github("ropensci/rnaturalearthhires")
install.packages(
  "rnaturalearthhires",
  repos = "https://ropensci.r-universe.dev",
  type = "source"
)

install.packages("prettymapr")

library(prettymapr)
library(rnaturalearth)
library(rnaturalearthdata)
library(rnaturalearthhires)
library(treemapify)
library(patchwork)
library(galah)
library(readr)
library(tidyr)
library(dplyr)
library(stringr)
library(ggplot2)
library(purrr)
library(sf)
library(tibble)

library(austraits)    # functions for exploring a traits.build database, available at Github repo
library(traits.build) # additional functions for exploring a traits.build database, available at Github repo

source("https://raw.githubusercontent.com/traitecoevo/traits.build-book/master/data/extra_functions.R")

aus_traits <- austraits::get_versions() %>%
  dplyr::pull("doi") %>%
  dplyr::first()

aus_traits

austraits <- austraits::load_austraits(doi = aus_traits)
austraits
```

# data wrangling
importing data, combining my dataset with AusTraits, filtering by searched for species
```{r}
#importing my data
setwd("/Users/dylan/Desktop/FUni/honours/trait_analysis_honours")
master_data <- read.csv("data_searching_cleaned.csv")
master_data

#filtering by species we searched for
searched_for_data <- master_data %>% # filtering only populations we can search for
  filter(refindable == "Y")

#fixing taxonomic issues
searched_for_data <- searched_for_data %>%
  mutate(species_name = case_when(
    species_name == "Calamagrostis quadriseta" ~ "Deyeuxia quadriseta",
    species_name == "Sphaeropteris australis" ~ "Cyathea leichhardtiana",
    species_name == "Chenopodium trigonon" ~ "Einadia trigonos",
    TRUE ~ species_name))

#simplifying found status
searched_for_data <- searched_for_data %>%
  mutate(found_simplified = case_when( # simplify found or not to just yes or no
    found %in% c("Y", "?Y") ~"Yes",
    found %in% c("N", "?N") ~"No",
    found %in% c("?", "F", "") ~"?",
    TRUE ~ found)) %>%
  mutate(found_simplified_withF = case_when( # simplify found or not to just yes or no
    found %in% c("Y", "?Y") ~"Yes",
    found %in% c("N", "?N") ~"No",
    found %in% c("?", "") ~"?",
    found %in% c("F") ~ "Failed search",
    TRUE ~ found)) %>%
  mutate(found_simplified_number = case_when(
    found_simplified %in% c("Yes") ~1,
    found_simplified %in% c("No") ~0,
    found_simplified %in% c("?") ~NA))

#removing "?"
searched_for_data_YorN <- searched_for_data %>%
  filter(found_simplified != "?")

#to be able to choose the order of these response variables in my graph
searched_for_data$found_factor <- factor(searched_for_data$found, levels = c("Y", "?Y", "N", "?N", "F", "?", "")) 

#TRIMMING AUSTRAITS DATA to just my species
austraits_searching <- 
  austraits$traits %>%
  filter(taxon_name %in% c(searched_for_data$species_name)) %>% # filter by only for species I searched for
  left_join(searched_for_data %>% # add re-found or not column
              select(species_name, found), by = c("taxon_name" = "species_name")) %>%
  mutate(found_simplified = case_when( # simplify found or not to just yes or no
    found %in% c("Y", "?Y") ~"Yes",
    found %in% c("N", "?N") ~"No",
    found %in% c("?", "F", "") ~"?",
    TRUE ~ found)) %>%
  mutate(found_simplified_number = case_when(
    found_simplified %in% c("Yes") ~1,
    found_simplified %in% c("No") ~0,
    found_simplified %in% c("?") ~NA))

#filtering out "?" found responses

austraits_searching_foundYorN <- 
  austraits_searching %>%
  filter(found_simplified != "?")

n_distinct(austraits_searching_foundYorN$taxon_name) # checking number of species that I've filtered to
setdiff(searched_for_data$species_name, austraits_searching_foundYorN$taxon_name) # checking missing species - should be good now (the 7 missing ones had "?" found value)



# adding woodiness data
woodiness_data <- austraits_searching %>%
  filter(trait_name == "woodiness_detailed")

# filtering out conflicting values
woodiness_data_cleaned <- 
  woodiness_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) == 1) %>% # filtering out species with conflicting values
  distinct(taxon_name, .keep_all = TRUE) %>%
  ungroup()

#filtering out "?" for graphs and stats - didnt do originally as adding woodiness data to main dataset, to wanted to have it for all
woodiness_data_cleaned <- 
  woodiness_data_cleaned %>%
  filter(found_simplified != "?")

#ADDING WOODINESS DATA TO MAIN DATA TABLE
searched_for_data <- 
  searched_for_data %>%
  select(-any_of("woodiness")) %>% #removing column if it already exists (to avoid duplicates)
  left_join(
    woodiness_data_cleaned %>%
      select(taxon_name, value), by = c("species_name" = "taxon_name")) %>%
  rename(woodiness = value) # renaming column to "woodiness"


# searching a row in data frame
searched_for_data %>%
  filter(species_name == "Flagellaria indica") %>% # change species name and / or trait name
  select(species_name, woodiness) # what columns i want to view
```

# AusTrait queries
```{r}
austraits::lookup_trait(database = austraits, term = "dispersal") # use to look up traits: place search term after term = "..."

filter(austraits$traits,
       taxon_name %in% c("Gonocarpus tetragynus"),
       trait_name == "woodiness_detailed") # use to look up species in AusTraits

austraits %>%  # check values of a given trait
  pluck("traits") %>%
  filter(trait_name == "dispersal_appendage") %>% #CHANGE TRAIT NAME HERE
  pull("value") %>% 
  unique()

# checking which species are missing values
austraits_searching_foundYorN %>%
  distinct(taxon_name) %>%
  anti_join(
    austraits_searching_foundYorN %>%
      filter(trait_name == "seedbank_longevity") %>%
      distinct(taxon_name),
    by = "taxon_name"
  )

# checking what species have a given trait entry
austraits_searching_foundYorN %>%
  filter(trait_name == "serotiny") %>% # enter trait here
  arrange(taxon_name)
```

# found Y/N graphs
```{r}
found_graph <- 
  ggplot(searched_for_data,
         aes (x = found_simplified_withF, fill = found_simplified_withF)
         ) +
    geom_bar() +
    geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, fontface = "bold") +
    theme_minimal() +
   scale_fill_manual(
      values = c("Yes" = "orange", "No" = "darkblue", "Failed search" = "brown", "?" = "grey")) +
    scale_x_discrete(labels = c("?" = "Unknown")) +
    theme(legend.position = "none") +
    labs(x = "Found?",
         y = "Species") 

found_graph
```


# year, decade
plotting re-findability by year
```{r}
year_plot <- 
  ggplot(searched_for_data, aes (x=sampled_year, fill = found_simplified_withF)) + # bar plot with samples collected by year
    geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "Failed search" = "brown", "?" = "grey")
  ) +
  labs(
    x = "Year",
    y = "Collections",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(searched_for_data$sampled_year), max(searched_for_data$sampled_year), by = 2),
                     labels = function(x) ifelse(x %% 4 == 0, x, ""))
year_plot

#without unknowns
year_plot_YorN <- 
  ggplot(searched_for_data_YorN, aes (x=sampled_year, fill = found_simplified_withF)) + # bar plot with samples collected by year
    geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    x = "Year",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_continuous(breaks = seq(min(searched_for_data$sampled_year), max(searched_for_data$sampled_year), by = 2),
                     labels = function(x) ifelse(x %% 4 == 0, x, ""))
year_plot_YorN

# by decade
searched_for_data <-
  searched_for_data %>%
  mutate(sampled_decade = floor(sampled_year / 10) * 10)

decade_plot <- 
  ggplot(searched_for_data, aes (x=factor(sampled_decade), fill = found_simplified_withF)) + # factor() makes sampled_decade categorical
    geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "Failed search" = "brown", "?" = "grey")
  ) +
  labs(
    x = "Decade",
    y = "Collections",
    fill = "Found?") + 
  theme_minimal() + 
  scale_x_discrete(labels = function(x) paste0(x, "s")) # making it "1990s", not "1990"
decade_plot
```
# age of sample
re-findability by age of sample - PROBABLY DONT NEED THIS AS IT'S JUST SAMPLE YEAR BUT FLIPPED
```{r}
ggplot(searched_for_data, aes (x = age_of_sample_years, fill = found_simplified_withF)) +
  geom_bar() +
  scale_fill_manual(values = c("Yes" = "orange", "No" = "darkblue", "Failed search" = "brown", "?" = "grey"),
                    labels = c("?" = "Unknown")
  ) +
  labs(
    x = "Age of sample",
    y = "Count",
    fill = "Found?",
    title = "Re-findability by age of sample") +
  theme_minimal()

# test
glm(as.numeric(found_simplified_number) ~ age_of_sample_years,
    family = binomial(link = "logit"),
    data = searched_for_data_YorN) %>%
  summary()
```

# height data wrangling
height data curation
```{r}
# Extract height data
height_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "plant_height", life_stage != "seedling") %>% #filtering only plant height as a trait, as every trait is in the same column. also filtering out values that came from seedlings
  mutate(value=as.numeric(value)) # changing plant height data to numerical (IMPORANT!!!)

table(height_data$unit) ##### for checking ######

#NOTE::::there are also min and mean values, but not using these
max_height_data <- height_data %>% 
  filter(value_type == "maximum")

#add a column with mean max plant height to data table
max_height_data_cleaned <- max_height_data %>% 
  group_by(taxon_name) %>%
  mutate(average_height = mean(value, na.rm = TRUE)) %>%
  distinct(taxon_name, .keep_all = TRUE)
```

# height violin plots
```{r}
height_plot <- 
  ggplot(max_height_data_cleaned, aes(x = average_height, y = found_simplified, fill = found_simplified)) +
  geom_violin() +
  theme_minimal() +
  labs(
    title = "Re-findability by plant height (n = 77)",
    x = "Average height (m)", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey"))
height_plot

#log transformed
log_height_plot <- 
  ggplot(max_height_data_cleaned, aes(x = average_height, y = found_simplified, fill = found_simplified)) +
  geom_violin() +
  theme_minimal() +
  labs(
    title = "Re-findability by plant height (n = 77)",
    x = "Average height (log transformed m)", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_log10() +
  theme(legend.position = "none")
log_height_plot
```

# height trend line, stats
```{r}
#dot plot with trend line
max_height_data_searched %>%
  filter(found_simplified_number != "?") %>%
  ggplot(aes(x = average_height, y = found_simplified_number)) +
  geom_point() +
  scale_x_log10() +
 geom_smooth(method = "glm", method.args = list(family = "binomial")) + # adding trend line
  theme_minimal() +
  labs(
    title = "Re-findability by plant height (n = 75)",
    x = "Average height", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("1" = "orange", "0" = "darkblue"))

summary(
  max_height_data_searched %>%
  na.omit() %>%
  glm(as.numeric(found_simplified_number) ~ log10(average_height),
      family = binomial(link = "logit"),
      data = .)
  )
```

# woodiness
filtering and graphing woodiness data
```{r}
#NOTE - some species have different values depending on source, when filtering, the most common is selected
#Eleocharis sphacelata - one “herbaceous woody_root”, 5 “herbaceous”
#gonocarpus tetragynus - one “herbaceous woody_base”, 4 “herbaceous”
#Flagellaria indica - has both herbaceous (2) and woody (1)
#Monoxtaxis linifolia - "woody base" or "herbaceous woody base" (1 each) #
#Piper hederaceum, Opercularia aspera - herbaceous or herbaceous woody (1 each) #
#Senecio pinnatifolius - one “herbaceous semi-woody”, numerous “herbaceous)

#checking species with conflicting trait values 
woodiness_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) > 1) %>% #filtering for multiple trait values
  count(taxon_name, value) %>% #adding a count column
  arrange(taxon_name) # sorting by species


table(woodiness_data_cleaned$value) # checking

woodiness_plot <- 
  ggplot(woodiness_data_cleaned, aes(x = value, fill = found_simplified)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Re-findability by woodiness (n = 88)",
    x = "Woodiness", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_discrete(
    labels = c("herbaceous" = "Herbaceous", "woody" = "Woody"))
woodiness_plot

woodiness_table <- table(woodiness_data_cleaned$value, woodiness_data_cleaned$found_simplified) # creating a contingency table
chisq.test(woodiness_table) # chi squared test
chisq.test(woodiness_table)$residuals # gives a value for each value interaction

```
# life history
annual vs perennial
```{r}
life_history_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "life_history")

table(life_history_data$value)

#checking species with conflicting trait values 
life_history_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) > 1) %>% #filtering for multiple trait values
  distinct(taxon_name, value) %>% #removing duplicates
  arrange(taxon_name) # sorting by species

# checking which species have values that are not annual or perennial
life_history_data %>%
  group_by(taxon_name) %>%
  filter(!value %in% c("annual", "perennial")) %>% #filtering for multiple trait values
  arrange(taxon_name) # sorting by species


#TODO work out what to do with weird values e.g. annual perennial
life_history_data_cleaned <- 
  life_history_data %>%
  group_by(taxon_name) %>%
  mutate(value = case_when( # simplifying
    value %in% c("short_lived_perennial", "perennial") ~ "perennial",
    value %in% c("annual biennial", "annual") ~ "annual",
    TRUE ~ value)) %>%
  filter(n_distinct(value) == 1) %>%
  slice(1) %>% # keeping only the first entry for each species
  arrange(taxon_name) # sorting by species




#        ALTERNATIVE CODE to include conflicting values (not using)
#life_history_data_cleaned <- 
#  life_history_data %>%
#  group_by(taxon_name) %>%
#  mutate(value = case_when( # simplifying
#    value %in% c("short_lived_perennial", "perennial") ~ "perennial",
#    value %in% c("annual biennial", "annual") ~ "annual",
#    TRUE ~ value)) %>%
#  mutate(value = case_when(
#    n_distinct(value) > 1 ~ "conflicting",
#    TRUE ~ value
#  )) %>%
#  slice(1) %>% # keeping only the first entry for each species
#  arrange(taxon_name) # sorting by species




life_history_plot <-
  ggplot(life_history_data_cleaned, aes(x = value, fill = found_simplified)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Re-findability by life history (n = 81)",
    x = "Life history", 
    y = "Species",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_discrete(
    labels = c("annual" = "Annual", "perennial" = "Perennial"))
life_history_plot

#chi squared test
life_history_table <- table(life_history_data_cleaned$found_simplified, life_history_data_cleaned$value)
chisq.test(life_history_table) # chi squared test
chisq.test(life_history_table)$residuals # gives a value for each value interaction

```

# resprouting
```{r}
resprouting_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "resprouting_capacity") %>%
  arrange(taxon_name)
resprouting_data

table(resprouting_data$value)

# checking which species are missing values
austraits_searching_foundYorN %>%
  distinct(taxon_name) %>%
  anti_join(
    austraits_searching_foundYorN %>%
      filter(trait_name == "vegetative_reproduction_ability") %>%
      distinct(taxon_name),
    by = "taxon_name"
  )

#checking which species have conflicting values
resprouting_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) > 1) %>% #filtering for multiple trait values
  distinct(taxon_name, value) %>% #removing duplicates
  arrange(taxon_name) # sorting by species

#making species with conflicting values partial re-sprouting
resprouting_data_cleaned <- resprouting_data %>%
  group_by(taxon_name) %>%
  mutate(value = if (n_distinct(value) > 1) "partial_resprouting" else unique(value)) %>%
  slice(1) %>%
  ungroup() %>%
  arrange(taxon_name)


# plot
resprouting_plot <-
  ggplot(resprouting_data_cleaned, aes(x = value, fill = found_simplified)) +
  geom_bar(position = "stack") + # think about position = "dodge" for all
  theme_minimal() +
  labs(
    title = "Re-findability by resprouting capability \n(n = 80)",
    x = "Resprouting capability", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
scale_x_discrete(
  labels = c("fire_killed" = "Killed by fire", "partial_resprouting" = "Partial\nresprouting", "resprouts" = "Resprouter"))
resprouting_plot

#chi squared test
resprouting_table <- table(resprouting_data_cleaned$found_simplified, resprouting_data_cleaned$value)
chisq.test(resprouting_table) # chi squared test
chisq.test(resprouting_table)$residuals # gives a value for each value interaction
```

# seedbank
NOTE sample size only \~30
```{r}
seedbank_data <- austraits_searching_foundYorN %>%
  filter(trait_name == "seedbank_location") %>%
  mutate(value_simplified = case_when(
    value %in% c("soil_seedbank", "canopy_seedbank") ~"seedbank",
    TRUE ~ value)) %>%
  arrange(taxon_name)

#checking which species have conflicting values
seedbank_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) > 1) %>% #filtering for multiple trait values
  distinct(taxon_name, value) %>% #removing duplicates
  arrange(taxon_name)# sorting by species 

seedbank_data_cleaned <- 
  seedbank_data %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) == 1) %>% # removing species with conflicting values
  slice(1) %>% # keeping only the first entry for each species
  arrange(taxon_name) # sorting by species


#plot it
seedbank_plot <-
  ggplot(seedbank_data_cleaned, aes(x = value, fill = found_simplified)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Re-findability by seedbank location (n = 33)",
    x = "Seedbank type", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_discrete(
    labels = c("canopy_seedbank" = "Canopy\nseedbank", "soil_seedbank" = "Soil\nseedbank", "none" = "None"))
seedbank_plot


#chi squared test
seedbank_table <- table(seedbank_data_cleaned$found_simplified, seedbank_data_cleaned$value)
chisq.test(seedbank_table) # chi squared test
chisq.test(seedbank_table)$residuals # gives a value for each value interaction


# simplified
seedbank_simplified_plot <-
  ggplot(seedbank_data_cleaned, aes(x = value_simplified, fill = found_simplified)) +
  geom_bar() +
  theme_minimal() +
  labs(
    title = "Re-findability by seedbank presence (n = 33)",
    x = "Seedbank presence", 
    y = "Species count",
    fill = "Found?") +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  scale_x_discrete(
    labels = c("seedbank" = "Present", "none" = "Absent"))
seedbank_simplified_plot

seedbank_simplified_table <- table(seedbank_data_cleaned$found_simplified, seedbank_data_cleaned$value_simplified)
chisq.test(seedbank_simplified_table) # chi squared test
chisq.test(seedbank_simplified_table)$residuals # gives a value for each value interaction
```
# search type (how many)
searched individually or alongside other species
```{r}
individual_or_group_search_plot <- ggplot(searched_for_data_YorN, aes(x = individual_or_multiple_search, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by search type (n = 95)",
    x = "Was the target species searched for\nexclusively or alongside others?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("I" = "Exclusively", "M" = "Alongside others"))
individual_or_group_search_plot

#chi squared 
individual_or_multiple_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$individual_or_multiple_search)
chisq.test(individual_or_multiple_table) # chi squared test
chisq.test(individual_or_multiple_table)$residuals # gives a value for each value interaction
```
# familiarity
searcher's have seen species before?
```{r}
seen_before_plot <- 
  ggplot(searched_for_data_YorN, aes(x = volunteers_seen_before, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by species familiarity (n = 95)",
    x = "Had the searchers seen\nthe species before?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("N" = "No", "Y" = "Yes"))
seen_before_plot

#chi squared 
volunteers_seen_before_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$volunteers_seen_before)
chisq.test(volunteers_seen_before_table) # chi squared test
chisq.test(volunteers_seen_before_table)$residuals # gives a value for each value interaction
```

# galah wrangling
galah and range edge data manipulation
```{r}
galah_config(atlas = "ALA", #selecting ALA as database
             email = "d.petrozzi@student.unsw.edu.au")

#creating data frame with all ALA records of my species (have hashed out as don't need to do every time)
ALA_data <-
  galah_call() %>%
  identify(searched_for_data$species_name) %>%
  collect()


#USE TO CHECK INDIVIDUAL SPECIES
#ALA_data %>%
#  filter(scientificName == "Baloskion tetraphyllum") %>%
#  filter(!is.na(decimalLatitude)) %>% # removing NAs
#  pull(decimalLatitude) %>% # extracting numeric vector (converting single column data frame to a numeric vector)
#  ecdf() %>%
#  { .(-34.13) * 100 } # curly brackets to immediately call the ecdf function with the Royal latitude (SHOULD CONFIRM ROYAL LATITUDE)


#pulling a list of species names as they appear in ALA
ALA_taxa_data <- search_taxa(searched_for_data$species_name) %>%
  select(search_term, scientific_name) %>% # choosing columns to remain
  distinct(search_term, scientific_name) %>%
  collect()

#and then joining to existing dataframe
searched_for_data <- 
  searched_for_data %>%
  select(-any_of("scientific_name_ALA")) %>% #removing column if it already exists (to avoid duplicates)
  left_join(ALA_taxa_data, by = c("species_name" = "search_term")) %>%
  rename(scientific_name_ALA = scientific_name) # renaming column to scientific_name_ALA


#adding range quantile column - NOTE: NEED TO CHANGE NAMES OF A FEW SPECIES AS DIFFER BETWEEN MY SHEET AND ALA ( i think i have done this now)
searched_for_data <-
  searched_for_data %>%
  select(-any_of("latitude_percentile")) %>% # removing column if it's already there to avoid duplication
  left_join(
    ALA_data %>%
      filter(occurrenceStatus == "PRESENT") %>%
      filter(!is.na(decimalLatitude)) %>% # removing NAs
      group_by(scientificName) %>%
      summarise(latitude_percentile = ecdf(decimalLatitude)(-34.13)*100), #empirical cumulative distribution functions (ecdf)
    by = c("scientific_name_ALA" = "scientificName")
  )
```

# range edge graphs
```{r}
searched_for_data %>%
  filter(found_simplified_number != "?") %>%
  ggplot(aes(x = latitude_percentile, y = found_simplified_number)) +
  geom_point() +
 geom_smooth(method = "glm", method.args = list(family = "binomial")) + # adding trend line
  theme_minimal() +
  labs(
    title = "Re-findability by range location",
    x = "Range quantile", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("1" = "orange", "0" = "darkblue"))

range_plot <- searched_for_data %>%
  filter(found_simplified_number != "?") %>%
  ggplot(aes(x = latitude_percentile, y = found_simplified, fill = found_simplified)) +
  geom_violin() +
  theme_minimal() +
  labs(
    title = "Re-findability by latitude quantile (n = 95)",
    x = "Latitude quantile", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) +
  theme(legend.position = "none") +
  scale_x_continuous(
    breaks = c(0, 25, 50, 75, 100),
    labels = c("0\n(South\nrange\nedge)", "25", "50", "75", "100\n(North\nrange\nedge)"))
range_plot

glm(as.numeric(found_simplified_number) ~ latitude_percentile,
    family = binomial(link = "logit"),
    data = searched_for_data) %>% # work our why na.omit doesn't work
  summary()
```
# unique genera in royal
unique genera in Royal analysis and plot
```{r}
#ALTERNATIVE APPROACH (probably better) using Thomas' curated species list (avoids homonym error also)
royal_master_list <- 
  read.csv("royal_master_list.csv") %>%
  mutate(genus = word(accepted_name, 1)) # adding genera column

unique_genera_data <- 
  royal_master_list %>%
  group_by(genus) %>%
  filter (n() == 1) %>%
  ungroup() %>%
  filter(accepted_name %in% searched_for_data$species_name)

searched_for_data <-
  searched_for_data %>%
  mutate(unique_genera_in_royal = species_name %in% unique_genera_data$accepted_name)



##PLOT AND STATS

unique_in_royal_plot <- 
  searched_for_data %>%
  filter(found_simplified != "?") %>%
  ggplot(aes(x = unique_genera_in_royal, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")
  ) +
  labs(
    title = "Re-findability by uniqueness (n = 95)",
    x = "Is the species the only one\nof its genera in Royal?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(labels = c("FALSE" = "No", "TRUE" = "Yes"))
unique_in_royal_plot

unique_in_royal_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$unique_genera_in_royal)
chisq.test(unique_in_royal_table) # chi squared test
chisq.test(unique_in_royal_table)$residuals # gives a value for each value interaction
```


importing park borders for monotypic data analysis (ALSO REDUNDANT AS USING THOMAS' PARK SPECIES LIST)
```{r}
Royal_outline <- st_read("royal_map/Royal_outline.shp") # importing polygon (shapefile)

# checking if the boundary box aligns with original shape file (no need to create simplified one if we are using a boundary box as it sidesteps the issue of vertices as it is just a rectangle)
Royal_outline %>%
  { ggplot() +
      geom_sf(data = ., fill = "lightblue", alpha = 0.5) +
      geom_sf(data = st_as_sfc(st_bbox(.)), fill = NA, color = "red", size = 1) +
      ggtitle("polygon (shapefile) vs bounding box")
  }

# adding column with genera
searched_for_data <-
  searched_for_data %>%
  mutate(genus = word(scientific_name_ALA, 1))

#searching for all species with genera that we searched for that exist in Royal - SEE WHY SOME GENERA DON'T WORK (error - homonym)
genera_searched_Royal <- 
  galah_call() %>%
  identify(searched_for_data$genus) %>%
  galah_bbox(st_bbox(Royal_outline_simplified_vertices)) %>%
  collect()

# filtering for species which are monotypic of their genus (IN ROYAL)
genera_searched_Royal %>%
  distinct(scientificName) %>%
  mutate(genus = word(scientificName, 1)) %>%
  group_by(genus) %>%
  filter (n() == 1) %>%
  ungroup() %>%
  filter(scientificName %in% searched_for_data$scientific_name_ALA)
```

reducing vertices and creating a boundary box (DONT NEED TO USE IF DOING SPATIAL DATA SCRAPING WITH galah_bbox)
```{r}
#counting number of vertices (needs to be under 500 for galah_polygon)
Royal_outline_simplified_vertices %>%
  st_geometry() %>%
  lapply(st_coordinates) %>%
  lapply(nrow) %>%
  unlist() %>%
  sum()

# create a function to count vertices
count_vertices <- function(geom) {
  geom %>%
    lapply(st_coordinates) %>%
    lapply(nrow) %>%
    unlist() %>%
    sum()
}

# creating values for max vertices and how much we adjust tolerance each time we change it
tolerance <- 0.01

# running the repeating code to reduce vertices
repeat {
  Royal_outline_simplified_vertices <- 
    Royal_outline %>%
    st_simplify(preserveTopology = TRUE, dTolerance = tolerance) # main repeating segment

  total_vertices <- Royal_outline_simplified_vertices %>% # creating a value for the new number of vertices
    st_geometry() %>%
    count_vertices()

  if (total_vertices <= 500) { # checking if vertices in new shapefile is less than 500
    break # ending repeating formula if true
  }

  tolerance <- tolerance + 0.01 # adding to tolerance each repear
}

# plotting the original and simplified vertices map on top of each other to make sure it worked
ggplot() +
  geom_sf(data = Royal_outline, fill = 'lightblue') +
  geom_sf(data = Royal_outline_simplified_vertices, fill = NA, color = 'red') +
  ggtitle("Original")

# checking if the BOUNDARY BOX aligns with simplified vertices shapefile (using boundary box for analysis as the shapefile is too complex to run)
Royal_outline_simplified_vertices %>%
  { ggplot() +
      geom_sf(data = ., fill = "lightblue", alpha = 0.5) +
      geom_sf(data = st_as_sfc(st_bbox(.)), fill = NA, color = "red", size = 1) +
      ggtitle("Polygon with Bounding Box")
  }

```

# fertile
found with fertile material?
```{r}
found_fertile_plot <- 
  searched_for_data_YorN %>%
  filter(found_simplified == "Yes") %>%
  ggplot(aes(x = fertile_material_present, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by fertile material prescence",
    x = "was fertile material present? (in those found)",
    y = "count",
    fill = "Found?") + 
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_discrete(labels = c("no", "yes"))
found_fertile_plot

found_sterile_table <- table(
  searched_for_data_YorN$found_simplified[
    searched_for_data_YorN$found_simplified == "Yes"],
  searched_for_data_YorN$fertile_material_present[
    searched_for_data_YorN$found_simplified == "Yes"]
)
chisq.test(found_sterile_table) # chi squared test
chisq.test(found_sterile_table)$residuals # gives a value for each value interaction
```

# associated species
```{r}
associated_species_plot <- 
  ggplot(searched_for_data_YorN, aes(x = associated_species_known, fill = found_simplified)) + 
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by associated species\n(n = 95)",
    x = "Were associated species known?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("N" = "No", "Y" = "Yes"))
associated_species_plot

#chi squared 
associated_species_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$associated_species_known)
chisq.test(associated_species_table) # chi squared test
chisq.test(associated_species_table)$residuals # gives a value for each value interaction
```

# search type (shape)
linear or area search?
```{r}
linear_or_area_plot <- 
  ggplot(searched_for_data_YorN, aes(x = linear_or_area_search, fill = found_simplified)) + 
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by search shape (n = 95)",
    x = "Was a linear or area-based\nsearch conducted?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(labels = c("A" = "Area", "L" = "Linear"))
linear_or_area_plot

#chi squared 
linear_or_area_table <- table(searched_for_data_YorN$found_simplified, searched_for_data_YorN$linear_or_area_search)
chisq.test(linear_or_area_table) # chi squared test
chisq.test(linear_or_area_table)$residuals # gives a value for each value interaction
```

# flowering time
searching during flowering time?
```{r}
flowering_time_plot <- 
  searched_for_data_YorN %>%
  filter(searched_in_flowering_time != "#N/A") %>%
   ggplot(aes(x = searched_in_flowering_time, fill = found_simplified)) +
  geom_bar() +
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue")
  ) +
  labs(
    title = "Re-findability by search timing (n = 71)",
    x = "Was search conducted during\nflowering time?",
    y = "Species count",
    fill = "Found?") + 
  theme_minimal() +
  scale_x_discrete(
    labels = c("N" = "No", "Y" = "Yes"))
flowering_time_plot

flowering_time_table <- 
  searched_for_data_YorN %>%
  filter(searched_in_flowering_time != "#N/A") %>%
  with(table(found_simplified, searched_in_flowering_time))
chisq.test(flowering_time_table) # chi squared test
chisq.test(flowering_time_table)$residuals # gives a value for each value interaction

#checking what species were found with fertile material outside flowering time
searched_for_data %>%
  filter(searched_in_flowering_time == "N", 
         fertile_material_present == "Y",
         found_simplified == "Yes") %>%
  select(species_name, searched_in_flowering_time, fertile_material_present, found)
```

# range edge x woodiness
(need to check code and asumptions here)
```{r}
searched_for_data %>%
  filter(found == "?")

rangexwoodiness_plot <- 
  searched_for_data_YorN %>%
  filter(!is.na(woodiness)) %>%
  ggplot(aes(x = latitude_percentile, 
             y = woodiness, 
             fill = found_simplified)) +
  geom_boxplot(outlier.shape = NA) + # makes the outlier normal shaped
  theme_minimal() +
  labs(
    title = "Re-findability by range and woodiness",
    x = "range value", 
    y = "woodiness",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey")) + 
  geom_jitter(size = 0.5, alpha = 1,
              position = position_dodge(width = 0.75))
rangexwoodiness_plot
aov(found_simplified_number ~ woodiness * latitude_percentile, data = searched_for_data_YorN) %>%
  summary()
```


# map wrangling
```{r}
library(sf)
library(rmapshaper)
library(ggspatial)

#map!!
Royal_map <- st_read("royal_map/Royal_outline.shp") %>%
  st_transform(crs = 4326) %>% # assigning WGS 84 coordinate system
  st_make_valid()

Royal_roads <- st_read("royal_map/Royal_roads.gpx", layer = "tracks") %>% # reading road file
  st_transform(crs = 4326) %>%
  st_make_valid() %>%
  { rmapshaper::ms_clip(target = ., clip = Royal_map[1]) } # limiting extent of roads to borders of Royal map

searched_for_data <-
  searched_for_data %>%
  separate(search_area_coordinates, into = c("search_area_lat", "search_area_long"), sep = ", ", convert = TRUE) # splitting lat and long into separate columns for mapping

searched_for_data_sf <- searched_for_data %>%
  st_as_sf(coords = c("search_area_long", "search_area_lat"), crs = 4326) %>%
  st_transform(crs = st_crs(Royal_map))
```


# building the maps
```{r}
#ALL longunobserved
royal_all_long_unobserved <- read.csv("royal_long_unobserved_all.csv")

royal_all_long_unobserved_sf <- royal_all_long_unobserved %>%
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) %>%
  st_transform(crs = st_crs(Royal_map))

all_long_unobserved_map <-
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = royal_all_long_unobserved_sf, aes(color = establishment_means), size = 1, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  +
   scale_color_manual(values = c(
     "native" = "darkblue",
     "introduced" = "orange"),
     labels = c(
       "native" = "Native",
       "introduced" = "Introduced"
     )) +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  ) +
  labs(
    colour = "Establishment means"
  )


# fig 1
long_unobserved_native_by_year_map <-
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = subset(royal_all_long_unobserved_sf, establishment_means == "native"), aes(color = year), size = 1, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  +
   scale_color_gradientn(
    name = "Year last recorded",
    colors = c("yellow", "orange", "purple4"),
    breaks = c(seq(1870, 1990, by = 20)),
    limits = c(1870, 1990),
    na.value = "purple4",
    guide = guide_colourbar(
    barheight = unit(7, "cm"),
    barwidth = unit(0.5, "cm"))) +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  ) +
  labs(
    colour = "Establishment means"
  )
long_unobserved_native_by_year_map


#map searched by year
searched_map_by_year <- 
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = searched_for_data_sf, aes(color = sampled_year), size = 1.5, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  +
   scale_color_gradient(
    name = "year sampled",
    low = "yellow",
    high = "purple") +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  )

#map searched by found # FIG 6
searched_map_by_found <- 
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = searched_for_data_sf, aes(color = found_simplified_withF), size = 1.5, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  +
   scale_color_manual(
    name = "Search outcome",
    values = c(
    "Yes" = "orange",
    "No" = "darkblue",
    "?" = "grey",
    "Failed search" = "brown"),
    labels = c(
      "?" = "Unknown (1)",
      "Yes" = "Found (27)",
      "No" = "Not found (68)",
      "Failed search" = "Search not conducted (4)"
    )) +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  )


all_long_unobserved_map
searched_map_by_year + searched_map_by_found
```

# maps with inset Aus
```{r}
royal_bbox <- st_as_sfc(st_bbox(Royal_outline)) # making Royal_outline into a bounding box compatible with ggspatial (sf polygon)
inset_aus_map <- inset_element(ggplot() +
  geom_sf(data = aus_map_simple, fill = "#D8E7D0", color = "grey50") +
  geom_sf(data = royal_coords, color = "red", size = 1, shape = 0, stroke = 0.7) +
  theme_void() +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.3)
  ), 
    left = 0.7,   # horizontal placement (0 = left, 1 = right)
    bottom = -0.4, # vertical placement (0 = bottom, 1 = top)
    right = 0.98,  # width of inset area
    top = 0.98     # height of inset area
)
inset_aus_map

long_unobserved_native_by_year_map + inset_aus_map

searched_map_by_found + inset_aus_map
```



```{r}
#long unobserved native status AND year last recorded
long_unobserved_native_year_map <-
  ggplot() + 
  geom_sf(mapping = aes(), data = Royal_map , stat = "sf", fill = "#C5DDC2", color = "grey50", alpha = 0.8) +
  geom_sf(mapping = aes(), data = Royal_roads, stat = "sf") +
  geom_sf(data = royal_all_long_unobserved_sf, aes(shape = establishment_means, color = year), size = 1.5, alpha = 1) +
  annotation_scale(location = "br", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm"), line_width = 0.5, height = unit(0.2, "cm")) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() + 
    annotation_north_arrow(location = "tr",      # top right
                         which_north = "true", # or "grid" north
                         style = north_arrow_fancy_orienteering(), # other styles available
                         height = unit(1, "cm"),
                         width = unit(1, "cm"),
                         pad_x = unit(0.5, "cm"),
                         pad_y = unit(0.5, "cm"))  + 
  scale_color_gradient(
    low = "orange",              
    high = "darkblue") +
  scale_shape_manual(values = c(
    "native" = 16,
    "introduced" = 2
  )) +
  theme(
    legend.position = "right",
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    axis.ticks.length = unit(0, "pt"),
    plot.title = element_text(hjust = 0, size = 10),
    plot.margin=unit(c(.0,.0,.0,.0),"cm")
  ) +
  labs(
    shape = "Establishment means",
    color = "Year sampled"
  )

long_unobserved_native_year_map
```


# ALL PLOTS 
```{r}

# creating a theme for figure format
figure_theme <- (function(p) { 
    p + theme(legend.position = "none", 
              plot.title = element_text(face = "bold", size = 10))})

#combining plots

trait_plots <-
  list(
  woodiness_plot,
    resprouting_plot,
    seedbank_simplified_plot,
    log_height_plot, 
    range_plot
    ) %>%
    lapply(figure_theme) %>%
    wrap_plots(ncol = 2, guides = "collect") +
    plot_annotation(tag_levels = 'a') +
    theme(legend.position = "right")

methodology_plots <-
  list( 
    individual_or_group_search_plot, 
    seen_before_plot,
    unique_in_royal_plot,
    associated_species_plot, 
    linear_or_area_plot,
    flowering_time_plot
    ) %>%
    lapply(figure_theme) %>%
    wrap_plots(ncol = 2, guides = "collect") + # guide refers to things like legends
    plot_annotation(tag_levels = 'a') +  #adding labels a, b, c...
    theme(legend.position = "right")     # Set overall legend position



# METHODS

long_unobserved_native_by_year_map
fig_1 <- long_unobserved_native_by_year_map + inset_aus_map # fig 1
  ggsave("plots/fig_1.png", plot = fig_1, width = 6, height = 6, dpi = 300)


family_all_species_plot # fig 2
  ggsave("plots/fig_2.png", plot = family_all_species_plot, width = 8, height = 6, dpi = 300)

range_edge_species_maps # fig 3
  ggsave("plots/fig_3.png", plot = range_edge_species_maps, width = 8, height = 6, dpi = 300)
north_range_maps
south_range_maps

curation_treemap # fig S1
  ggsave("plots/fig_S1.png", plot = curation_treemap, width = 8, height = 5, dpi = 300)


# RESULTS 

found_graph 

searched_map_by_year
searched_map_by_found
fig_6 <-searched_map_by_found + inset_aus_map # fig 6
  ggsave("plots/fig_6.png", plot = fig_6, width = 6, height = 6, dpi = 300)

year_plot_YorN # fig 7
  ggsave("plots/fig_7.png", plot = year_plot_YorN, width = 8, height = 6, dpi = 300)
year_plot
decade_plot

trait_plots # fig 8
  ggsave("plots/fig_8.png", plot = trait_plots, width = 8, height = 6, dpi = 300)
methodology_plots # fig 9
  ggsave("plots/fig_9.png", plot = methodology_plots, width = 8, height = 6, dpi = 300)

abundance_density_plot # fig 10
  ggsave("plots/fig_10.png", plot = abundance_density_plot, width = 8, height = 6, dpi = 300)
rangexwoodiness_plot

gaia_final_map # fig 11
  ggsave("plots/fig_11.png", plot = gaia_final_map, width = 3, height = 5, dpi = 300)
```

# theoretical extinction
```{r}
theoretical_extinction <- 
  tibble(
    "year" = c(1975, 1990, 2025),
    "persistence_probability" = c(1, 0.9, 0.3)
  )

theoretical_extinction

ggplot(theoretical_extinction, 
       aes(x = year, 
           y = persistence_probability)) +
  geom_line() +
  geom_ribbon(aes(ymin = persistence_probability - 0.1, ymax = persistence_probability + 0.1), alpha = 0.3) +
  scale_x_continuous(
    limits = c(1975, 2025),
    breaks = seq(min(theoretical_extinction$year), max(theoretical_extinction$year), by = 5)
  ) +
  theme_minimal() +
  labs(
    y = "p(X|t)"
  )
```


# tree map
```{r}
#fixing names as they differ on ALA
master_data <- 
  master_data %>%
  mutate(species_name = case_when(
    species_name == "Calamagrostis quadriseta" ~ "Deyeuxia quadriseta",
    species_name == "Sphaeropteris australis" ~ "Cyathea leichhardtiana",
    species_name == "Chenopodium trigonon" ~ "Einadia trigonos",
    species_name == "Cyanothamnus rigens" ~ "Boronia rigens",
    species_name == "Setaria distans" ~ "Paspalidium distans",
    species_name == "Microsorum pustulatum" ~ "Zealandia pustulata", # multiple synonyms, but the record links to this species
    species_name == "Genoplesium nudiscapum" ~ "Corunastylis nudiscapa",
    species_name == "Genoplesium woollsii" ~ "Corunastylis woollsii",
    species_name == "Lastreopsis decomposita" ~ "Parapolystichum calanthum", # AusTraits uses this name for the one trait record, as opposed to the accepted Lastreopsis calantha
    species_name == "Stolonochloa pygmaea" ~ "Panicum pygmaeum",
    species_name == "Zostera capricorni" ~ "Zostera muelleri subsp. capricorni",
    species_name == "Gaudium parvifolium" ~ "Leptospermum parvifolium",
    TRUE ~ species_name))

#filtering AusTraits to all long unobserved
austraits_long_unobserved <- 
  austraits$traits %>%
  filter(taxon_name %in% c(master_data$species_name)) %>%
  left_join(master_data %>% # add re-found or not column
              select(species_name, taxonomy_issue, location_issue, refindable), by = c("taxon_name" = "species_name"))
# and then to woodiness
woodiness_long_unobserved <-
  austraits_long_unobserved %>%
  filter(trait_name == "woodiness_detailed") %>%
  filter(value %in% c("woody", "herbaceous")) %>%
  group_by(taxon_name) %>%
  filter(n_distinct(value) == 1) %>%
  ungroup() %>%
  distinct(taxon_name, .keep_all = TRUE) %>%
  rename(woodiness = value) %>% # renaming column to "woodiness" 
  select(taxon_name, woodiness, taxonomy_issue, location_issue) %>%
  mutate(refindable = case_when(
    location_issue == "Y" & taxonomy_issue == "Y" ~ "Location and taxonomy issue",
    location_issue == "Y" & taxonomy_issue == "N" ~ "Location issue",
    location_issue == "N" & taxonomy_issue == "Y" ~ "Taxonomy issue",
    location_issue == "N" & taxonomy_issue == "N" ~ "No issues"
  ))


n_distinct(woodiness_long_unobserved$taxon_name)

woodiness_treemap <- 
  woodiness_long_unobserved %>%
  count(woodiness, refindable) %>%
  mutate(woodiness = factor(woodiness))


# plot
curation_treemap <- 
  ggplot(woodiness_treemap, 
         aes(area = n, 
             fill = refindable, 
             subgroup = woodiness, 
             label = n)) +
    geom_treemap(color = "black", size = 2) +  
    geom_treemap_subgroup_border(color = "black", size = 2) + 
    geom_treemap_text(
      aes(label = n),
      place = "centre",
      size = 8,
      colour = "white",
      grow = FALSE,
      fontface = "bold"
    ) +
    geom_treemap_subgroup_text(
    place = "centre",
    colour = "black", 
    size = 12, 
    fontface = "bold"
    ) +
    labs(fill = "Re-findable?") +
    theme(legend.position = "right") +
  scale_fill_manual(
    values = c("Location and taxonomy issue" = "darkblue",
               "Location issue" = "grey",
               "Taxonomy issue" = "brown",
               "No issues" = "orange"),
    labels = c("Location and taxonomy issue" = "No (Location and taxonomy issue)",
               "Location issue" = "No (Location issue)",
               "Taxonomy issue" = "No (Taxonomy issue)",
               "No issues" = "Yes"),
     breaks = c("No issues", "Taxonomy issue", "Location issue", "Location and taxonomy issue"))

curation_treemap
```
old treemap code (not using)
```{r}
install.packages("treemap")
library(treemap)

treemap(woodiness_treemap,
        index = c("woodiness", "n", "refindable"),
        vSize = "n",
        type = "categorical",
        vColor = "refindable",
        title = "woodiness by sample issues",
        fontsize.labels = c(12, 8, 0), # removing woodiness labels from graph
        bg.labels = c("transparent")
)
```


# australia range maps
```{r}
# checking range edge species
searched_for_data %>%
  arrange(latitude_percentile) %>%
  select(species_name, latitude_percentile)

# importing australia map
aus_map <- ne_countries(scale = "medium", country = "Australia", returnclass = "sf") # Australia map
aus_states <- ne_states(country = "Australia", returnclass = "sf") # state boundaries
royal_coords <- st_sf(
              name = "Royal National Park", 
              geometry = st_sfc(st_point(c(151.07, -34.13))), 
              crs = 4326)

aus_map_simple <- ne_countries(scale = "small", country = "Australia", returnclass = "sf")

# creating a theme for maps format
map_theme <- 
  function(m) { 
    m + theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "italic", , size = 9))}


## NORTHERN RANGE edges

#Juncus pallidus	96.1713173			
#Leptospermum continentale	96.8019113			
#Gahnia radula	97.6496396	


# JUNCUS PALLIDUS map
juncus_pallidus_coords <-
  galah_call() %>%
  identify("Juncus pallidus") %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

juncus_pallidus_map <-
  ggplot(data = aus_map) +
  geom_sf(fill = "#C5DDC2", 
          color = "grey50", 
          alpha = 0.8) +
  geom_sf(data = aus_states, 
          fill = NA, 
          color = "gray30") +
  geom_sf(data = juncus_pallidus_coords, 
          aes(color = "Observations"), 
          size = 1.3) +
  geom_sf(data = royal_coords, 
          aes(color = "Royal National Park"), 
          size = 2) +
  scale_color_manual(
    values = c(
      "Royal National Park" = "blue4", 
      "Observations" = "darkorange"),
    name = NULL) +
  coord_sf(
    xlim = c(110, 157),   # Longitude range of Australia
    ylim = c(-45, -9),   # Latitude range of Australia
    expand = FALSE
  ) + 
  theme_minimal() +
  labs(
    title = "Juncus pallidus"
  )



# Leptospermum continentale map
leptospermum_continentale_coords <-
  galah_call() %>%
  identify("Leptospermum continentale") %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

leptospermum_continentale_map <- 
  ggplot(data = aus_map) +
  geom_sf(fill = "#C5DDC2", 
          color = "grey50", 
          alpha = 0.8) +
  geom_sf(data = aus_states, 
          fill = NA, 
          color = "gray30") +
  geom_sf(data = leptospermum_continentale_coords, 
          aes(color = "Observations"), 
          size = 1.3) +
  geom_sf(data = royal_coords, 
          aes(color = "Royal National Park"), 
          size = 2) +
  scale_color_manual(
    values = c(
      "Royal National Park" = "blue4", 
      "Observations" = "darkorange"),
    name = NULL) +
  coord_sf(
    xlim = c(110, 157),   # Longitude range of Australia
    ylim = c(-45, -9),   # Latitude range of Australia
    expand = FALSE
  ) + 
  theme_minimal() +
  labs(
    title = "Leptospermum continentale"
  )


# Gahnia radula
gahnia_radula_coords <-
  galah_call() %>%
  identify("Gahnia radula") %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

gahnia_radula_map <- 
 ggplot(data = aus_map) +
  geom_sf(fill = "#C5DDC2", 
          color = "grey50", 
          alpha = 0.8) +
  geom_sf(data = aus_states, 
          fill = NA, 
          color = "gray30") +
  geom_sf(data = gahnia_radula_coords, 
          aes(color = "Observations"), 
          size = 1.3) +
  geom_sf(data = royal_coords, 
          aes(color = "Royal National Park"), 
          size = 2) +
  scale_color_manual(
    values = c(
      "Royal National Park" = "blue4", 
      "Observations" = "darkorange"),
    name = NULL) +
  coord_sf(
    xlim = c(110, 157),   # Longitude range of Australia
    ylim = c(-45, -9),   # Latitude range of Australia
    expand = FALSE
  ) + 
  theme_minimal() +
  labs(
    title = "Gahnia radula"
  )




## SOUTHERN RANGE EDGEs

#Flagellaria indica	0.1828525			
#Cyperus cyperoides	0.8771930			
#Sacciolepis indica	0.9615385	


# Flagellaria indica map
flagellaria_indica_coords <-
  galah_call() %>%
  identify("Flagellaria indica") %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

flagellaria_indica_map <-
  ggplot(data = aus_map) +
  geom_sf(fill = "#C5DDC2", 
          color = "grey50", 
          alpha = 0.8) +
  geom_sf(data = aus_states, 
          fill = NA, 
          color = "gray30") +
  geom_sf(data = flagellaria_indica_coords, 
          aes(color = "Observations"), 
          size = 1.3) +
  geom_sf(data = royal_coords, 
          aes(color = "Royal National Park"), 
          size = 2) +
  scale_color_manual(
    values = c(
      "Royal National Park" = "blue4", 
      "Observations" = "darkorange"),
    name = NULL) +
  coord_sf(
    xlim = c(110, 157),   # Longitude range of Australia
    ylim = c(-45, -9),   # Latitude range of Australia
    expand = FALSE
  ) + 
  theme_minimal() +
  labs(
    title = "Flagellaria indica"
  )



# Cyperus cyperoides map
cyperus_cyperoides_coords <-
  galah_call() %>%
  identify("Cyperus cyperoides") %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

cyperus_cyperoides_map <-
   ggplot(data = aus_map) +
  geom_sf(fill = "#C5DDC2", 
          color = "grey50", 
          alpha = 0.8) +
  geom_sf(data = aus_states, 
          fill = NA, 
          color = "gray30") +
  geom_sf(data = cyperus_cyperoides_coords, 
          aes(color = "Observations"), 
          size = 1.3) +
  geom_sf(data = royal_coords, 
          aes(color = "Royal National Park"), 
          size = 2) +
  scale_color_manual(
    values = c(
      "Royal National Park" = "blue4", 
      "Observations" = "darkorange"),
    name = NULL) +
  coord_sf(
    xlim = c(110, 157),   # Longitude range of Australia
    ylim = c(-45, -9),   # Latitude range of Australia
    expand = FALSE
  ) + 
  theme_minimal() +
  labs(
    title = "Cyperus cyperoides"
  )



# Sacciolepis indica map
sacciolepis_indica_coords <-
  galah_call() %>%
  identify("Sacciolepis indica") %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)

sacciolepis_indica_map <-
  ggplot(data = aus_map) +
  geom_sf(fill = "#C5DDC2", 
          color = "grey50", 
          alpha = 0.8) +
  geom_sf(data = aus_states, 
          fill = NA, 
          color = "gray30") +
  geom_sf(data = sacciolepis_indica_coords, 
          aes(color = "Observations"), 
          size = 1.3) +
  geom_sf(data = royal_coords, 
          aes(color = "Royal National Park"), 
          size = 2) +
  scale_color_manual(
    values = c(
      "Royal National Park" = "blue4", 
      "Observations" = "darkorange"),
    name = NULL) +
  coord_sf(
    xlim = c(110, 157),   # Longitude range of Australia
    ylim = c(-45, -9),   # Latitude range of Australia
    expand = FALSE
  ) + 
  theme_minimal() +
  labs(
    title = "Sacciolepis indica"
  )


range_edge_species_maps <- 
  list(juncus_pallidus_map, leptospermum_continentale_map, gahnia_radula_map, 
     flagellaria_indica_map, cyperus_cyperoides_map, sacciolepis_indica_map) %>%
  lapply(map_theme) %>%
  wrap_plots(ncol = 3, guides = "collect") + 
  plot_annotation(tag_levels = 'a') &
  theme(legend.position = "bottom")

north_range_maps <- 
  list(juncus_pallidus_map, leptospermum_continentale_map, gahnia_radula_map) %>%
  lapply(map_theme) %>%
  wrap_plots(ncol = 3, guides = "collect") + 
  plot_annotation(tag_levels = 'a') &
  theme(legend.position = "bottom")

south_range_maps <- 
  list(flagellaria_indica_map, cyperus_cyperoides_map, sacciolepis_indica_map) %>%
  lapply(map_theme) %>%
  wrap_plots(ncol = 3, guides = "collect") +
  plot_annotation(tag_levels = 'a') &
  theme(legend.position = "bottom")
```

# range map template
```{r}

plot_range_map <- function(species) {

  galah_call() %>%
  identify(species) %>%
  geolocate(aus_map_simple) %>%
  collect() %>%
  filter(!is.na(decimalLatitude), !is.na(decimalLongitude)) %>%
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) %>%
  {ggplot(data = aus_map) +
    geom_sf(fill = "#C5DDC2", 
            color = "grey50", 
            alpha = 0.8) +
    geom_sf(data = aus_states, 
            fill = NA, 
            color = "gray30") +
    geom_sf(data = ., 
            aes(color = "Observations"), 
            size = 1.3) +
    geom_sf(data = royal_coords, 
            aes(color = "Royal National Park"), 
            size = 2) +
    scale_color_manual(
      values = c(
        "Royal National Park" = "blue4", 
        "Observations" = "darkorange"),
      name = NULL) +
    coord_sf(
      xlim = c(110, 157),   # Longitude range of Australia
      ylim = c(-45, -9),   # Latitude range of Australia
      expand = FALSE
    ) + 
    theme_minimal() +
      labs(
        title = species
      )
    }  
}



# checking range edge species
searched_for_data %>%
  arrange(latitude_percentile) %>%
  select(species_name, latitude_percentile)

plot_range_map("Poa poiformis")
```

# resprouting x woodiness
```{r}

```

# family count of all long unobserved
```{r}
family_all_species_data <-
  master_data %>%
    count(family) %>%
    mutate(family = if_else(n < 5, "Other families", family)) %>%
    group_by(family) %>% 
    summarise(n = sum(n), .groups = "drop") %>%
    mutate(family = reorder(family, n)) %>%
    mutate(life_form = case_when(
      family %in% c("Cyperaceae", "Poaceae", "Juncaceae", "Restionaceae") ~ "Graminoids",
      family %in% c("Myrtaceae", "Fabaceae", "Asteraceae", "Rhamnaceae", "Proteaceae", "Polygonaceae",
                      "Plantaginaceae", "Lauraceae", "Ericaceae", "Araliaceae") ~ "Eudicots",
      family %in% c("Orchidaceae", "Asparagaceae") ~ "Non-graminoid monocots",
      TRUE ~ "Other families"
    ))

family_all_species_plot <- 
  ggplot() + 
    geom_bar(data = filter(family_all_species_data, family != "Other families"),
             aes(x = family, y = n, fill = life_form),
             stat = "identity") +
    geom_bar(data = filter(family_all_species_data, family == "Other families"),
             aes(x = family, y = n, fill = life_form),
             stat = "identity",
             show.legend = FALSE) +
    coord_flip() +
    geom_text(
      data = family_all_species_data,
      aes(x = family, y = n, label = n),
      hjust = -0.5, 
      size = 3, 
      fontface = "bold") +
    theme_classic() + 
    theme() +
    scale_fill_manual(
      values = c("Graminoids" = "brown",
                 "Eudicots" = "orange",
                 "Non-graminoid monocots" = "darkblue"
    )) +
    labs(y = "Species count",
         fill = "Life form",
         x = "Family") +
  scale_y_continuous(
    limits = c(0, 120)
  )

family_all_species_plot
```

# seed longevity
```{r}
seed_longevity_data <- 
  austraits_searching_foundYorN %>%
    filter(trait_name == "seedbank_longevity") %>%
    arrange(taxon_name) %>%
  mutate(
    min = as.numeric(str_extract(value, "\\d+\\.?\\d*")),
    max = as.numeric(sapply(str_extract_all(value, "\\d+\\.?\\d*"), tail, 1))
  ) %>%
  group_by(taxon_name) %>%
  slice_max(order_by = max, with_ties = FALSE) %>%
  ungroup() %>%
  select(taxon_name, value, min, max, basis_of_record, found, found_simplified, found_simplified_number)

seed_longevity_plot <- 
  ggplot(seed_longevity_data, aes(x = max, y = found_simplified, fill = found_simplified)) +
  geom_violin() +
  theme_minimal() +
  labs(
    title = "Re-findability by seed longevity (n = 63)",
    x = "seedbank longevity (years)", 
    y = "Found?",
    fill = "Found?") + 
  scale_fill_manual(
    values = c("Yes" = "orange", "No" = "darkblue", "?" = "grey"))
seed_longevity_plot

summary(
  seed_longevity_data %>%
  na.omit() %>%
  glm(as.numeric(found_simplified_number) ~ max,
      family = binomial(link = "logit"),
      data = .)
  )
```
# abundance when found
```{r}
abundance_data <- searched_for_data %>%
  mutate(individuals_found_cleaned = case_when(
         individuals_found %in% c("") ~ NA,
         TRUE ~ individuals_found),
         individuals_found_cleaned = parse_number(individuals_found_cleaned)) %>% # pulling only numbers
  mutate(growth_form = case_when(
    quadrat_size_m2 %in% c(64) ~ "Herb",
    quadrat_size_m2 %in% c(144) ~ "Shrub",
    quadrat_size_m2 %in% c(400) ~ "Tree")) %>%
  filter(found_simplified == "Yes") %>%
  mutate(abundance_density = individuals_found_cleaned / quadrat_size_m2,
        abundance_density = case_when(
      individuals_found == "many" ~ 0.7, # manually assign abundance density to "many"
      TRUE ~ abundance_density)) %>%
  mutate(abundance_density = as.numeric(abundance_density)) %>%
  select(species_name, individuals_found, individuals_found_cleaned, found_simplified, quadrat_size_m2, abundance_density, growth_form)

# absolute abundance (misleading)
abundance_found_plot <- 
  ggplot(na.omit(abundance_data), aes(x = individuals_found_cleaned, fill = growth_form)) + ##### THIS IS MISSING "MANY" values ####
  geom_histogram(binwidth = 10, 
                 color = "black") + # changing bin width and colours
  stat_bin(binwidth = 10, 
           geom = "text", 
           aes(label = ifelse(..count.. == 0, "", ..count..)), vjust = -0.5) + # adding labels to each bar > 0
  theme_minimal() +
  labs(
    title = "Abundance of re-found species (n = 26)",
    x = "Individuals found within quadrat") +
  scale_x_continuous(
    breaks = seq(0, 60, by = 10),
    labels = c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", ">59")
  ) +
  scale_fill_manual(values = c("Herb" = "orange", "Shrub" = "brown", "Tree" = "darkblue"))
abundance_found_plot

# checking searches where over 10 indivduals were found
abundance_data %>%
  arrange(individuals_found_cleaned) %>%
  filter(individuals_found_cleaned > 10)


# abundance per area
abundance_density_plot <- 
  ggplot(
    filter(abundance_data, !is.na(abundance_density)), 
    aes(x = abundance_density*10, fill = growth_form)) +
  geom_histogram(binwidth = 1, 
                 color = "black",
                 linewidth = 0.4) +
  geom_text(
    data = filter(abundance_data, !is.na(abundance_density)),
    aes(x = abundance_density * 10, label = ifelse(..count.. == 0, "", ..count..)),
    stat = "bin",
    binwidth = 1,
    color = "black",
    vjust = -0.5,
    inherit.aes = FALSE
  ) +
  theme_classic() +
  labs(
    x = bquote("Aboveground stems per 10m"^2),
    y = "Species",
    fill = "Growth form") +
  scale_x_continuous(breaks = seq(0, 7, by = 1),
                     labels = c("<1", "1 - <2", "2 - <3", "3 - <4", "4 - <5", "5 - <6", "6 - <7", "≥7")) +
  scale_fill_manual(values = c("Herb" = "brown", "Shrub" = "orange", "Tree" = "darkblue"))
abundance_density_plot
```
# charismatic bias
in long unobserved
```{r}
# checking family counts #

native_royal_species_list <- royal_master_list %>%
  filter(establishment_means == "native") %>%
  mutate(long_unobserved = case_when(accepted_name %in% master_data$species_name ~ "long_unobserved", 
                                     TRUE ~ "N")) %>%
  mutate(uncharismatic = case_when(family %in% c("Cyperaceae", "Poaceae", "Juncaceae", "Restionaceae") ~ "uncharismatic",
                                   TRUE ~ "N"))

charisma_table <- table(native_royal_species_list$long_unobserved, native_royal_species_list$uncharismatic)
charisma_table
chisq.test(charisma_table)
chisq.test(charisma_table)$residuals

ggplot(native_royal_species_list, aes (x = long_unobserved, fill = uncharismatic)) +
  geom_bar()

```

# presence / absence
```{r}
ggplot(ALA_data, aes(x = occurrenceStatus)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_minimal()
```

# GAIA path
(ChatGPT made this fully)
```{r}
gaia_gpx <- st_read("sphaeropteris-australis.gpx", layer = "tracks")  # GPX tracks

# -------------------------------
# 2️⃣ Transform to a common CRS (Web Mercator for OSM)
# -------------------------------
Royal_map_osm <- st_transform(Royal_map, 3857)
Royal_roads_osm <- st_transform(Royal_roads, 3857)
gaia_gpx_osm <- st_transform(gaia_gpx, 3857)

# -------------------------------
# 3️⃣ Compute bounding boxes
# -------------------------------
gaia_bbox_numeric <- st_bbox(gaia_gpx_osm)           # numeric bbox
gaia_bbox_osm <- st_as_sfc(gaia_bbox_numeric)       # polygon for plotting
Royal_bbox <- st_bbox(Royal_map_osm)

# -------------------------------
# 4️⃣ Main GPX map
# -------------------------------
gaia_path_map <- ggplot() +
  # Base map
  annotation_map_tile(type = "osm", zoom = 18) +
  
  # Polygon map and roads
  geom_sf(data = Royal_map_osm, fill = NA, color = "grey50", alpha = 0.8) +
  geom_sf(data = Royal_roads_osm, color = "grey30") +
  
  # GPX track
  geom_sf(data = gaia_gpx_osm, color = "red", linewidth = 0.7) +
  
  # Scale and north arrow
  annotation_scale(location = "bl", width_hint = 0.5, pad_y = unit(1.4, "cm"), pad_x = unit(.3, "cm")) +
  annotation_north_arrow(location = "tr", which_north = "true", style = north_arrow_fancy_orienteering(),
                         height = unit(1, "cm"), width = unit(1, "cm"), pad_x = unit(0.5, "cm"), pad_y = unit(0.5, "cm")) +
  
  # Coordinate limits around GPX with buffer
  coord_sf(
    xlim = c(gaia_bbox_numeric["xmin"], gaia_bbox_numeric["xmax"]),
    ylim = c(gaia_bbox_numeric["ymin"], gaia_bbox_numeric["ymax"]),
    expand = TRUE
  ) +
  
  # Clean theme
  theme_classic() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    plot.margin = unit(c(0,0,0,0), "cm")
  )

# -------------------------------
# 5️⃣ Inset Royal map
# -------------------------------
inset_royal_map <- inset_element(
  ggplot() +
    geom_sf(data = Royal_map_osm, fill = "#D8E7D0", color = "grey50") +
    geom_sf(data = gaia_bbox_osm, color = "red", linewidth = 0.9) +
    coord_sf(
      xlim = c(Royal_bbox["xmin"], Royal_bbox["xmax"]),
      ylim = c(Royal_bbox["ymin"], Royal_bbox["ymax"]),
      expand = TRUE
    ) +
    theme_void() +
    theme(
      panel.background = element_rect(fill = "white", color = NA),
      panel.border = element_rect(color = "black", fill = NA)
    ),
  left = 0.55, bottom = 0.35, right = 0.95, top = 1
)

# -------------------------------
# 6️⃣ Combine main map + inset
# -------------------------------
gaia_final_map <- gaia_path_map + inset_royal_map
gaia_final_map
```

# aus unphotographed species by family
```{r}
aus_unphotographed_species <- read.csv("aus_unphotographed_species.csv")
aus_unphotographed_species %>%
  count(Family, sort = TRUE)
```

```{r}
galah_call() %>%
  identify("Choretrum candollei") %>%
  galah_bbox(st_bbox(Royal_outline)) %>%
  collect() %>%
  arrange(eventDate)
```

```{r}
family_all_species_simplified_plot <- 
  ggplot() + 
    geom_bar(data = filter(family_all_species_data, family != "Other families"),
             aes(x = family, y = n, fill = life_form),
             stat = "identity") +
    geom_bar(data = filter(family_all_species_data, family == "Other families"),
             aes(x = family, y = n, fill = life_form),
             stat = "identity",
             show.legend = FALSE) +
    coord_flip() +
    geom_text(
      data = family_all_species_data,
      aes(x = family, y = n, label = n),
      hjust = -0.5, 
      size = 3, 
      fontface = "bold") +
    theme_classic() + 
    theme(legend.position = "none") +
    scale_fill_manual(
      values = c("Graminoids" = "orange",
                 "Eudicots" = "grey40",
                 "Non-graminoid monocots" = "grey40",
                 "Other families" = "grey40"
    )) +
    labs(y = "Species count",
         fill = "Life form",
         x = "Family") +
  scale_y_continuous(
    limits = c(0, 120)
  )

family_all_species_simplified_plot
  ggsave("charismatic_bias_families.png", plot = family_all_species_simplified_plot, width = 8, height = 6, dpi = 300)
```

